---
title: "Genetic diversity, inbreeding, and differentiation"
format:
  html:
    toc: true
---

This notebook contains the code used to make tables and figures relating to
genetic diversity, differentiation, and inbreeding estimates. This includes some
supplemental tests, such as the impacts of downsampling coverage and simulation
comparisons to the empirical data.

```{r}
#| label: load-packages
#| code-summary: Load up required packages and setup species and genome lists

set.seed(20)

library(data.table) # For reading in data while skipping columns
library(stringr) # For separating pop names into regions and sampling years
library(ggplot2) # For plotting
library(ggpubr) # For adding stats to graphs
library(cowplot) # For arranging plots into a multi-panel figure
library(ggtext) # For more control over axis text layout
library(dplyr) # For summarize function and piping
library(geosphere) # For including distances on F~ST~ plot
library(scales) # For some control of plot scales
library(png) # For reading in maps made in QGIS
library(grid) # For plotting maps made in QGIS with ggplot
library(dplyr) # For summary tables in groups
library(nlme) # For mixed effect models
library(broom.mixed) # For tidying mixed effect models
library(Hmisc) # For bootstrapped confidence intervals
library(ggsignif) # For significance levels in plots
library(bootstrap) # For jackknife

species <- c("picarus", "pargus", "csemiargus")
species_names <- c("Po. icarus", "Pl. argus", "Cy. semiargus")
genomes <- c("ilPolIcar1.1", "ilPleArgu1.3", "ilCyaSemi1.1")
```

## Heterozygosity (tables, statistics, and plots)

Read in heterozygosity results from genotype calls for the three species,
combine them into a table and save that table for the supplement.

### Individual heterozygosity table

```{r}
#| label: tbl-heterozygosity

# prep heterozygosity dataframe and fill it in with heterozygosity estimates for
# three focal species.

hz_all <- c()

for (i in c(1:length(species))) {
  s <- species[i]
  g <- genomes[i]
  hz <- fread(
    paste0(
      "../analyses/results/datasets/historical-",
      s,
      "-vg-notrans/analyses/heterozygosity/historical-",
      s,
      "-vg-notrans.",
      g,
      "_all.dp6_allsites-filts.filtered_mindp5-allsites-jointcall_allbal0.21-0.79.notrans.fmiss0.3.bcftools.het.tsv"
    ),
    header = TRUE,
    sep = "\t"
  )
  samples <- fread(
    paste0(
      "../analyses/results/datasets/historical-",
      s,
      "-vg-notrans/qc/historical-",
      s,
      "-vg-notrans.",
      g,
      "_all.sampleqc.tsv"
    ),
    header = TRUE,
    sep = "\t",
    select = c("sample", "population", "time")
  )
  hz <- merge(hz, samples, by = "sample")
  hz$species <- s
  hz_all <- rbind(hz_all, hz)
}

# general clean up of df - separate pop name to get year and region, make
# species, time period, and region a factor with clean names

hz_all$year <- str_sub(hz_all$population, -4, -1)
hz_all$region <- str_sub(hz_all$population, 1, -5)

hz_all$species <- factor(
  hz_all$species,
  levels = species,
  labels = species_names
)

hz_all$time <- factor(
  hz_all$time,
  levels = c("historical", "modern"),
  labels = c("Historical", "Modern")
)

hz_all$region <- factor(
  hz_all$region,
  levels = c("WSkane", "SWSkane", "ESkane", "SESkane", "NSmaland", "Oland"),
  labels = c(
    "W Skåne", "SW Skåne", "E Skåne", "SE Skåne", "N Småland", "Öland"
  )
)

# repalce region name for SE/SW Skåne as they are grouped with E/W Skåne

hz_all$region_group <- hz_all$region
hz_all[hz_all$region_group == "SW Skåne", ]$region_group <- "W Skåne"
hz_all[hz_all$region_group == "SE Skåne", ]$region_group <- "E Skåne"


hz_table <- hz_all[, c(
  "sample", "species", "time", "population", "hetperbp", "hetperalt"
)]

write.table(hz_table, "../tables/hz_ests.tsv",
  row.names = FALSE, quote = FALSE, sep = "\t"
)
```

### Heterozygosity as a response to time period statistics

Run statistical tests on heterozygosity - performing LMMs with heterozygosity as
a response to time period with population as a random factor. Also perform
jackknifes and bootstraps to get confidence intervals around the mean estimates.

```{r}
#| label: stats-heterozygosity

# get means and confidence intervals around mean heterozygosity using
# bootstrapping of samples

hz_means <- hz_all %>%
  group_by(species, population) %>%
  do(
    data.frame(
      rbind(
        Hmisc::smean.cl.boot(.$hetperbp, B = 1000)
      )
    )
  )

write.table(hz_means, "../tables/hz_means.tsv",
  row.names = FALSE, quote = FALSE, sep = "\t"
)

# prepare table for supplement with historical/modern heterozygosity comparisons

hz_stats <- c()

for (s in species_names) {
  hist <- mean(hz_all[hz_all$species == s & hz_all$time == "Historical", ]$hetperbp * 1000)
  mod <- mean(hz_all[hz_all$species == s & hz_all$time == "Modern", ]$hetperbp * 1000)
  # estimate ratio of modern to historical heterozygosity
  ratio <- mod/hist
  # run linear mixed model with heterozygosity as response to time period and pop as a random eff.
  model <- lme(hetperbp * 1000 ~ time, random = ~1|region_group, data = hz_all[hz_all$species == s, ])
  # extract model parameter estimates with tidy
  tidyed <- tidy(model, conf.int = TRUE, effects = "fixed")
  tidydf <- tidyed[2, c(3,4,6,7,8,9)]
  tidydf$hist.mean <- hist
  # perform jackknifing as another approach to confidence intervals around means
  jackknife <- quantile(jackknife(hz_all[hz_all$species == s & hz_all$time == "Historical", ]$hetperbp * 1000, mean)$jack.values, c(0.025, 0.975))
  tidydf$hist.low.jk.ci <- jackknife[1]
  tidydf$hist.high.jk.ci <- jackknife[2]
  tidydf$mod.mean <- mod
  jackknife <- quantile(jackknife(hz_all[hz_all$species == s & hz_all$time == "Modern", ]$hetperbp * 1000, mean)$jack.values, c(0.025, 0.975))
  tidydf$mod.low.jk.ci <- jackknife[1]
  tidydf$mod.high.jk.ci <- jackknife[2]
  tidydf$`mod/hist` <- ratio
  tidydf <- round(tidydf, 5)
  tidydf$species <- s
  tidydf <- tidydf[ , c("species", "hist.mean", "hist.low.jk.ci", "hist.high.jk.ci", "mod.mean", "mod.low.jk.ci", "mod.high.jk.ci", "mod/hist", "estimate", "std.error", "statistic", "conf.low", "conf.high", "p.value")]
  hz_stats <- rbind(hz_stats, tidydf)
}

write.table(hz_stats, "../tables/hz_stats.tsv",
  row.names = FALSE, quote = FALSE, sep = "\t"
)
```

### Heterozygosity boxplots

Plot the differences in heterozygosity as a boxplot, with points colored by
sampling region and time period on the y-axis.

```{r}
#| label: fig-het-box
#| fig-width: 6
#| fig-height: 2.5

hzbox <- ggplot(hz_all, aes(x = time, y = hetperbp * 1000)) +
  facet_grid(cols = vars(species)) +
  geom_boxplot(
    aes(x = time, y = hetperbp * 1000),
    inherit.aes = FALSE,
    outliers = FALSE
  ) +
  geom_point(
    aes(color = region, shape = time),
    position = position_jitterdodge(jitter.width = 0.05)
  ) +
  scale_shape_manual(values = c(17, 16), guide = "none") +
  ylab("<span style='font-size: 9pt'>Heterozygous sites per 1000bp</span><br><span style='font-size: 7pt'>(excl. transitions)</span>") +
  scale_y_continuous(
    labels = number_format(accuracy = 0.1),
    expand = expansion(mult = c(0.05, 0.20))
  ) +
  scale_color_manual(
    values = c("#244F6E", "#244F6E", "#658399", "#658399", "#8A2A3B", "#fecd03"),
    name = "Geographic Region"
  ) +
  geom_signif(
    data = hz_all[hz_all$species == "Cy. semiargus", ],
    comparisons = list(c("Historical", "Modern")),
    annotations="**",
    margin_top = 0,
    y_position = 3
  ) +
  theme_bw() +
  theme(
    legend.position = "none",
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_markdown(),
    strip.text.x = element_text(size = 10, face = "italic")
  )
```

## Inbreeding coefficients (tables, statistics, and plots)

Read in Froh results from genotype calls for the three species, format them and
combine them into a table and save that table for the supplement.

```{r}
#| label: froh-table

froh_all <- c()

for (i in c(1:length(species))) {
  s <- species[i]
  g <- genomes[i]
  froh <- fread(
    paste0(
      "../analyses/results/datasets/historical-",
      s,
      "-vg-notrans/plots/inbreeding/historical-",
      s,
      "-vg-notrans.",
      g,
      "_all.dp6_allsites-filts.filtered_mindp5-biallelic-jointcall_allbal0.21-0.79.notrans.fmiss0.3.bcftools.ind_froh.tsv"
    ),
    header = TRUE,
    sep = "\t",
    select = c("sample", "population", "time", "depth", "Froh")
  )
  froh$species <- s
  froh_all <- rbind(froh_all, froh)
}

froh_all <- froh_all[froh_all$sample %in% hz_all$sample, ]

froh_all$year <- str_sub(froh_all$population, -4, -1)
froh_all$region <- str_sub(froh_all$population, 1, -5)

froh_all$species <- factor(
  froh_all$species,
  levels = species,
  labels = c("Po. icarus", "Pl. argus", "Cy. semiargus")
)

froh_all$time <- factor(
  froh_all$time,
  levels = c("historical", "modern"),
  labels = c("Historical", "Modern")
)

froh_all$region <- factor(
  froh_all$region,
  levels = c("WSkane", "SWSkane", "ESkane", "SESkane", "NSmaland", "Oland"),
  labels = c("W Skåne", "SW Skåne", "E Skåne", "SE Skåne", "N Småland", "Öland")
)

froh_all$region_group <- froh_all$region
froh_all[froh_all$region_group == "SW Skåne", ]$region_group <- "W Skåne"
froh_all[froh_all$region_group == "SE Skåne", ]$region_group <- "E Skåne"

froh_table <- froh_all[, c(
  "sample", "species", "time", "population", "Froh"
)]

write.table(froh_table, "../tables/froh_ests.tsv",
  row.names = FALSE, quote = FALSE, sep = "\t"
)
```

### Inbreeding coefficient as a response to time period statistics

Run statistical tests on inbreed coeff - performing LMMs with Froh as
a response to time period with population as a random factor. Also perform
jackknifes and bootstraps to get confidence intervals around the mean estimates.
Largely the same code as heterozygosity above.

```{r}
#| label: stats-froh

froh_means <- froh_all %>%
  group_by(species, population) %>%
  do(
    data.frame(
      rbind(
        Hmisc::smean.cl.boot(.$Froh, B = 1000)
      )
    )
  )

write.table(froh_means, "../tables/froh_means.tsv",
  row.names = FALSE, quote = FALSE, sep = "\t"
)

froh_stats <- c()

for (s in species_names) {
  hist <- mean(froh_all[froh_all$species == s & froh_all$time == "Historical", ]$Froh)
  mod <- mean(froh_all[froh_all$species == s & froh_all$time == "Modern", ]$Froh)
  model <- lme(Froh ~ time, random = ~1|region_group, data = froh_all[froh_all$species == s, ])
  tidyed <- tidy(model, conf.int = TRUE, effects = "fixed")
  tidydf <- tidyed[2, c(3,4,6,7,8,9)]
  tidydf$hist.mean <- hist
  jackknife <- quantile(jackknife(froh_all[froh_all$species == s & froh_all$time == "Historical", ]$Froh, mean)$jack.values, c(0.025, 0.975))
  tidydf$hist.low.jk.ci <- jackknife[1]
  tidydf$hist.high.jk.ci <- jackknife[2]
  tidydf$mod.mean <- mod
  jackknife <- quantile(jackknife(froh_all[froh_all$species == s & froh_all$time == "Modern", ]$Froh, mean)$jack.values, c(0.025, 0.975))
  tidydf$mod.low.jk.ci <- jackknife[1]
  tidydf$mod.high.jk.ci <- jackknife[2]
  tidydf <- round(tidydf, 5)
  tidydf$species <- s
  tidydf <- tidydf[ , c(13, 7, 8, 9, 10, 11, 12, 1, 2, 3, 5, 6, 4)]
  froh_stats <- rbind(froh_stats, tidydf)
}

write.table(froh_stats, "../tables/froh_stats.tsv",
  row.names = FALSE, quote = FALSE, sep = "\t"
)
```

### Froh boxplots

Plot the differences in Froh as a boxplot, with points colored by
sampling region and time period on the y-axis. Same as heterozygosity.

```{r}
#| label: fig-froh-box
#| fig-width: 6
#| fig-height: 2.5

frohbox <- ggplot(froh_all, aes(x = time, y = Froh)) +
  facet_grid(cols = vars(species)) +
  geom_boxplot(
    aes(x = time, y = Froh),
    inherit.aes = FALSE,
    outliers = FALSE
  ) +
  geom_point(
    aes(color = region, shape = time),
    position = position_jitterdodge(jitter.width = 0.05)
  ) +
  scale_shape_manual(values = c(17, 16), guide = "none") +
  ylab("<span style='font-size: 9pt'>*F~RoH~* </span><span style='font-size: 7pt'>(*RoH* > 100kb)</span>") +
  scale_color_manual(
    values = c("#244F6E", "#244F6E", "#658399", "#658399", "#8A2A3B", "#fecd03"),
    name = "Geographic Region"
  ) +
  scale_y_continuous(
    labels = number_format(accuracy = 0.1),
    expand = expansion(mult = c(0.05, 0.20))
  ) +
  geom_signif(
    data = froh_all[froh_all$species == "Cy. semiargus", ],
    comparisons = list(c("Historical", "Modern")),
    annotations="**"
  ) +
  theme_bw() +
  theme(
    legend.position = "none",
    panel.grid = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_markdown(),
    strip.text.x = element_text(size = 10, face = "italic")
  )
```

### Froh adjusted heterozygosity for the supplement

Combine Froh and heterozygosity tables, then use Froh to adjust heterozygosity
with the equation het / (1 - Froh), which approximates the avg heterozygosity
outside RoH. Plot this as a faceted boxplot for the supplement.

```{r}
#| label: fig-roh-adj-het
#| fig-width: 5
#| fig-height: 7

adj_hz <- merge(froh_all, hz_all)

# adjust heterozygosity by Froh
adj_hz$adj_hz <- adj_hz$hetperbp * 1000 / (1 - adj_hz$Froh)

roh_het <- adj_hz[ , c("sample", "population", "time", "species", "region")]
roh_het2 <- roh_het
roh_het$het <- adj_hz$hetperbp * 1000
roh_het$type <- "Including RoH"
roh_het2$het <- adj_hz$adj_hz
roh_het2$type <- "Excluding RoH"
roh_het <- rbind(roh_het, roh_het2)

roh_het$type <- factor(roh_het$type,
levels = c("Including RoH", "Excluding RoH"))

ggplot(roh_het, aes(x = time, y = het)) +
  facet_grid(rows = vars(species), cols = vars(type), scales = "free_y") +
  geom_boxplot(
    aes(x = time, y = het),
    inherit.aes = FALSE,
    outliers = FALSE
  ) +
  geom_point(
    aes(color = region, shape = time),
    position = position_jitterdodge(jitter.width = 0.05)
  ) +
  scale_shape_manual(values = c(17, 16), guide = "none") +
  ylab("<span style='font-size: 9pt'>Heterozygous sites per 1000bp</span><br><span style='font-size: 7pt'>(excl. transitions)</span>") +
  scale_y_continuous(
    labels = number_format(accuracy = 0.1),
    expand = expansion(mult = c(0.05, 0.20))
  ) +
  scale_color_manual(
    values = c("#244F6E", "#244F6E", "#658399", "#658399", "#8A2A3B", "#fecd03"),
    name = "Geographic Region"
  ) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_markdown(),
    strip.text.y = element_text(size = 10, face = "italic")
  )

ggsave("../figures/suppfig-roh-adj-het.png", dpi = 300, create.dir = TRUE)
```

## Nucleotide diversity

Prepare a table of mean estimates of pi per population by taking means across
windows. Get confidence intervals by bootstrapping across windows.

```{r}
#| label: tbl-pi
#| eval: true

pi_all <- c()

# read in windowed pi results for population pairs
for (i in c(1:length(species))) {
  s <- species[i]
  g <- genomes[i]
  nucdiv <- fread(
    paste0(
      "../analyses/results/datasets/historical-",
      s,
      "-vg-notrans/analyses/pixy/historical-",
      s,
      "-vg-notrans.",
      g,
      "_all.dp6_allsites-filts.filtered_mindp5-allsites-jointcall_allbal0.21-0.79.notrans.fmiss0.3/pixy_pi.txt"
    ),
    header = TRUE,
    sep = "\t",
  )
  nucdiv$species <- s
  pi_all <- rbind(pi_all, nucdiv)
}

# Summarize pi results per pair, using bootstraps to get confidence intervals
pi_all <- pi_all %>%
  group_by(species, pop) %>%
  do(data.frame(rbind(Hmisc::smean.cl.boot(.$avg_pi, B = 10000))))

# Read in sampling info to group pops by
regions <- read.table("../config/pop_groups.tsv", header = TRUE)

# Merge sampling info with pi data
pi_all <- merge(pi_all, regions, by.x = "pop", by.y = "population")

pi_all$species <- factor(
  pi_all$species,
  levels = species,
  labels = species_names
)

write.table(pi_all,
  file = "../tables/pi_ests.tsv", sep = "\t",
  quote = FALSE, row.names = FALSE
)
```

## Population differention using *F*~ST~

Prepare a table of mean estimates of *F*~ST~ per pop pair by taking means
across windows. Get confidence intervals by bootstrapping across windows.

```{r}
#| label: tbl-fst
#| eval: true

fst_all <- c()

# read in windowed F_ST results for population pairs
for (i in c(1:length(species))) {
  s <- species[i]
  g <- genomes[i]
  fst <- fread(
    paste0(
      "../analyses/results/datasets/historical-",
      s,
      "-vg-notrans/analyses/pixy/historical-",
      s,
      "-vg-notrans.",
      g,
      "_all.dp6_allsites-filts.filtered_mindp5-allsites-jointcall_allbal0.21-0.79.notrans.fmiss0.3/pixy_fst.txt"
    ),
    header = TRUE,
    sep = "\t",
  )
  fst$species <- s
  fst_all <- rbind(fst_all, fst)
}

fst <- fst_all

# Set negative Fst to 0
fst[fst$avg_hudson_fst < 0, ]$avg_hudson_fst <- 0

# Summarize Fst results per pair, getting confidence intervals from bootstraps
fst <- fst %>%
  group_by(species, pop1, pop2) %>%
  do(data.frame(rbind(Hmisc::smean.cl.boot(.$avg_hudson_fst, B = 10000))))

# Read in sampling info, including coordinates
regions <- read.table("../config/pop_groups.tsv", header = TRUE)

# Merge sampling info with F_ST data
fst <- merge(fst, regions, by.x = "pop1", by.y = "population")
fst <- merge(fst, regions, by.x = "pop2", by.y = "population")

# Get rid of pairs within same time period, get pair 'name', and set time
# categories for pair
fst <- fst[abs(fst$year.x - fst$year.y) < 60, ]
fst$pair <- paste0(fst$region.x, "-", fst$region.y)
fst$time <- "placeholder"
fst[fst$year.x < 2000, ]$time <- "historical"
fst[fst$year.x > 2000, ]$time <- "modern"

# Calculate distance between population pairs
fst$dist <- as.vector(
  apply(
    fst[, c("lon.x", "lat.x", "lon.y", "lat.y")], 1, function(x) distm(c(x[1], x[2]), c(x[3], x[4]), fun = distGeo)
  )
)

# Remove pairs where populations contain single samples
fst <- fst[!fst$pop2 %in% c("ESkane1917", "WSkane1943"), ]
fst <- fst[!fst$pop1 %in% c("ESkane1917", "WSkane1943"), ]

# Set up variables for plotting
fst$species <- factor(
  fst$species,
  levels = c("picarus", "pargus", "csemiargus"),
  labels = c("Po. icarus", "Pl. argus", "Cy. semiargus")
)
fst$time <- factor(
  fst$time,
  levels = c("historical", "modern"),
  labels = c("Historical", "Modern")
)

write.table(fst,
  file = "../tables/fst_ests.tsv", sep = "\t",
  quote = FALSE, row.names = FALSE
)
```

Plot line plots of *F*~ST~ change over time.

```{r}
#| label: fig-fst-box
#| fig-width: 6
#| fig-height: 2.5
#| eval: true

fstplot <- ggplot(
  fst,
  aes(
    x = time,
    y = Mean,
    ymax = Upper,
    ymin = Lower,
    group = pair,
    col = dist / 1000,
    label = paste0(round(dist / 1000, 1), " km")
  )
) +
  geom_line(linewidth = 0.5) +
  geom_point(aes(shape = time), size = 2.5) +
  scale_shape_manual(values = c(17, 16), guide = "none") +
  facet_grid(cols = vars(species)) +
  theme_bw() +
  scale_color_gradient(
    low = "#55AFF4", high = "#132C44",
    name = "Distance (km)"
  ) +
  ylab("<span style='font-size: 9pt'>Hudson *F~ST~*</span>") +
  scale_y_continuous(
    labels = number_format(accuracy = 0.01)
  ) +
  guides(
    colour = guide_colourbar(position = "inside")
  ) +
  theme(
    # legend.background = element_rect(color = "black", size = 0.2),
    legend.title = element_text(size = 8),
    legend.text = element_text(size = 7),
    legend.key.size = unit(0.4, "cm"),
    legend.justification.inside = c(0.02, 0.9),
    axis.title.x = element_blank(),
    panel.grid = element_blank(),
    axis.title.y = element_markdown(),
    strip.text = element_text(face = "italic", size = 10)
  )
```

## Make Figure 1

Combine pictures of butterflies, sampling map made in QGIS, and the three plots
generated above into the final Figure 1.

```{r}
#| label: fig-diversity
#| fig-width: 10.5
#| fig-height: 7
#| eval: true

# initialize above generated plots as new objects for arranging
arrhz <- hzbox

arrfroh <- frohbox

arrfst <- fstplot

# read in map and make it a ggplot
map <- readPNG("../resources/gis/sampling_map/historical_map.png")
map <- rasterGrob(map, interpolate = TRUE)
map <- ggplot() +
  geom_blank() +
  annotation_custom(map, xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf) +
  theme(
    panel.background = element_rect(
      fill = "transparent", color = NA
    ),
    plot.background = element_rect(
      fill = "transparent", color = NA
    )
  )

# read in butterfly images and make them a ggplot
butterflies <- readPNG("../images/butterfly_species_photos.png")
butterflies <- rasterGrob(butterflies, interpolate = TRUE)
butterflies <- ggplot() +
  geom_blank() +
  annotation_custom(
    butterflies,
    xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf
  ) +
  theme(
    panel.background = element_rect(
      fill = "transparent", color = NA
    ),
    plot.background = element_rect(
      fill = "transparent", color = NA
    )
  )

# arrange butterfly photos and map into a left panel for the plot
map_photos <- plot_grid(
  butterflies, map,
  labels = c("A", "B"),
  nrow = 2,
  ncol = 1,
  rel_heights = c(0.4, 1)
)

# arrange map photos, plus three plots here into a final figure
plot_grid(map_photos,
  plot_grid(arrhz, arrfroh, arrfst, ncol = 1, nrow = 3, labels = c("C", "D", "E"), align = "v", axis = "lr"),
  ncol = 2, nrow = 1
) + bgcolor("white")

ggsave("../figures/figure1.png", dpi = 600, create.dir = TRUE)
ggsave("../figures/figure1.svg", dpi = 600, create.dir = TRUE)
```

## Coalescent simulations

We used coalescent simulations to verify that the empirical estimates from our
genomic data aligned with population genetic expectations. These were done with
Slendr using the msprime backend, manually running them with the Rscript found
at
[`analyses/slendr/butterfly-decline-predict.R`](../analyses/slendr/butterfly-decline-predict.R).
This code summarizes this data into distributions of simulated outcomes reported
in Table S10 and Figure S9.


```{r}
#| label: fig-sims
#| fig-width: 8
#| fig-height: 5

# prep the hz_means and froh_means objects, these weren't used in making any
# tables in the end, but are clean data frames for this
hz_means$year <- str_sub(hz_means$population, -4, -1)
hz_means$region <- str_sub(hz_means$population, 1, -5)
froh_means$year <- str_sub(froh_means$population, -4, -1)
froh_means$region <- str_sub(froh_means$population, 1, -5)


# prepare an object with the empirical results for the relevant pops
sims <- c()

sn <- species_names[3]
s <- species[3]
sdf <- hz_means[hz_means$species == sn, ]
regs <- unique(sdf$region)
for (r in regs) {
  if (r == "SESkane") {
  } else {
    histh <- sdf[sdf$region == r & sdf$year < 2000, ]$Mean
    histy <- sdf[sdf$region == r & sdf$year < 2000, ]$year
    modh <- sdf[sdf$region == r & sdf$year > 2000, ]$Mean
    mody <- sdf[sdf$region == r & sdf$year > 2000, ]$year
    region <- r
    row <- c(s, sn, region, mody, histy, modh/histh)
    sims <- rbind(sims, row)
  }
}

sims <- data.frame(sims)
colnames(sims) <- c("species", "species_name", "region", "modyear", "histyear", "hetratio")

# read in the simulation outputs from the sims folder for the relevant pops
# calculate the relevant ratios of heterozygosity over time, incl. adjusted het
simdata <- c()

for (i in 1:nrow(sims)) {
  base <- paste(
    sims[i, 1], tolower(sims[i, 3]), "2121", sims[i, 4], sims[i, 5], "exclfirst-anc100", sep = "-"
  )
  sim <- read.table(
    paste0(
      "../sims/exclfirst/",
      base,
      ".tsv"
    )
  )
  sim0 <- c()
  sim0$hetratio <- sim$V2
  sim0$Froh <- sim$V7
  sim0$adjHetRatio <- sim$V4
  sim0$species <- sims[i, 1]
  sim0$region <- sims[i, 3]
  sim0$time <- 0
  sim100 <- c()
  sim100$hetratio <- sim$V3
  sim100$Froh <- sim$V8
  sim100$adjHetRatio <- sim$V5
  sim100$species <- sims[i, 1]
  sim100$region <- sims[i, 3]
  sim100$time <- 100
  simdata <- rbind(
    simdata,
    data.frame(sim0)[ , c("species", "region", "hetratio", "Froh", "adjHetRatio", "time")],
    data.frame(sim100)[ , c("species", "region", "hetratio", "Froh", "adjHetRatio", "time")]
  )
}

sims[, "hetratio"] <- as.numeric(sims[, "hetratio"])
sims$species <- sims$species_name

froh_mod <- froh_means[froh_means$year > 2000, ]

simdata$species <- factor(
  simdata$species,
  levels = species,
  labels = species_names
)

# summarize quantiles for use in Table S10, which I manually make from
# what is printed here

hetratio <- simdata[simdata$time == 0, ] %>%
  group_by(region) %>%
  summarise(
    min = min(hetratio),
    q25 = quantile(hetratio, 0.25),
    q50 = quantile(hetratio, 0.50),
    q75 = quantile(hetratio, 0.75),
    max = max(hetratio)
  )

hetratio

modFroh <- simdata[simdata$time == 0, ] %>%
  group_by(region) %>%
  summarise(
    min = min(Froh),
    q25 = quantile(Froh, 0.25),
    q50 = quantile(Froh, 0.50),
    q75 = quantile(Froh, 0.75),
    max = max(Froh)
  )

modFroh

adjHetRatio <- simdata[simdata$time == 0, ] %>%
  group_by(region) %>%
  summarise(
    min = min(adjHetRatio),
    q25 = quantile(adjHetRatio, 0.25),
    q50 = quantile(adjHetRatio, 0.50),
    q75 = quantile(adjHetRatio, 0.75),
    max = max(adjHetRatio)
  )

adjHetRatio

simdata$time <- factor(
  simdata$time,
  levels = c(0, 100),
  labels = c("Sampled year", "100 years post sampling")
)

# drop SESkane from froh_mod object since it isn't in the sims
froh_mod <- froh_mod[froh_mod$region != "SESkane", ]

pd <- position_dodge(width = 0.9)

hetsim <- ggplot(sims, aes(x = region, y = hetratio)) +
  # plot distribution of simulation replicate values
  geom_violin(data = simdata, aes(linetype = time), position = pd) +
  # plot simulation median values
  stat_summary(data = simdata, aes(group = time), position = pd, fun = "median", color = "black", geom = "point") +
  # plot empirical estimates with some nudging to align it with the right time (sample year rather than 100 years post)
  geom_point(color = "red", size = 3, position = position_nudge(x = -0.9/4)) +
  ylab("Proportion retained heterozygosity") +
  theme_bw() +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    axis.title.x = element_blank()
  )

frohsim <- ggplot(froh_mod[froh_mod$species == "Cy. semiargus", ], aes(x = region, y = Mean)) +
  # plot distribution of simulation replicate values
  geom_violin(data = simdata, aes(y = Froh, linetype = time), position = pd) +
  # plot simulation median values
  stat_summary(data = simdata, aes(y = Froh, group = time), position = pd, fun = "median", color = "black", geom = "point") +
  # plot empirical estimates with some nudging to align it with the right time (sample year rather than 100 years post)
  geom_point(color = "red", size = 3, position = position_nudge(x = -0.9/4)) +
  ylab("Mean modern *F~RoH~*") +
  theme_bw() +
  theme(
    legend.position = "none",
    axis.title.y = element_markdown(),
    axis.title.x = element_blank()
  )

# combine into Figure S9
plot_grid(
  hetsim,
  frohsim,
  align = "h",
  ncol = 2
)

ggsave("../figures/suppfig-sims.png", dpi = 600)
```

## Compare with full depth in *Cy. semiargus*

To verify that the patterns we observed in the modern samples for heterozygosity
and Froh held even when samples were not subsampled to the coverage of
historical samples, we estimated heterozygosity, Froh, and Fst at full depth
using only the modern samples. This block plots the comparison depicted in
Figure S11.

```{r}
#| label: fig-het-fulldepth
#| fig-width: 10
#| fig-height: 5

# read in full depth with transition heterozygosity estimates for C. semiargus
s <- "csemiargus"
g <- "ilCyaSemi1.1"
hz <- fread(
  paste0(
    "../analyses/results/datasets/historical-",
    s,
    "-vg-trans-fulldp-modonly/analyses/heterozygosity/historical-",
    s,
    "-vg-trans-fulldp-modonly.",
    g,
    "_all_allsites-filts.filtered_mindp5-allsites-jointcall_allbal0.21-0.79.trans.fmiss0.3.bcftools.het.tsv"
  ),
  header = TRUE,
  sep = "\t"
)

# merge together original hz_all with new full depth hz. This makes hetperbp.x
# the downsampled het and hetperbp.y the full depth het
hz <- merge(hz_all, hz, by = "sample")

# read in sample qc table to get full depth value of each sample for plotting
fulldps <- fread("../analyses/results/datasets/historical-csemiargus-vg-trans-fulldp-modonly/qc/historical-csemiargus-vg-trans-fulldp-modonly.ilCyaSemi1.1_all.sampleqc.tsv",
  header = TRUE,
  sep = "\t"
)

# merge in the sample depth info to the hz object
hz <- merge(hz, fulldps, by = "sample")

# merge in downsampled dp Froh estimates to the hz object
hz <- merge(hz, froh_all[,c("sample", "Froh")], by = "sample")

# read in full depth with transitions Froh estimates
froh <- fread("../analyses/results/datasets/historical-csemiargus-vg-trans-fulldp-modonly/plots/inbreeding/historical-csemiargus-vg-trans-fulldp-modonly.ilCyaSemi1.1_all_allsites-filts.filtered_mindp5-biallelic-jointcall_allbal0.21-0.79.trans.fmiss0.3.bcftools.ind_froh.tsv")

# merge in full depth froh estimates, making Froh.x the downsampled and Froh.y
# the full depth estimate
hz <- merge(hz, froh[,c("sample", "Froh")], by = "sample")

# plot heterozygosity and Froh with downsampled on x and full dp on y
hz_full <- ggplot(hz, aes(x = hetperbp.x, y = hetperbp.y)) +
  geom_point(aes(color = `allsites-filts.depth.mean`)) +
  xlab("Heterozygous sites per bp (downsampled to 6x, no transitions)") +
  ylab("Heterozygous sites per bp (full depth, including transitions)") +
  geom_smooth(method = "lm", color = "black") +
  theme_classic() +
  labs(color = "Full seq. depth") +
  theme(
    legend.position = c(0.2, 0.7)
  )

f_full <- ggplot(hz, aes(x = Froh.x, y = Froh.y)) +
  geom_point(aes(color = `allsites-filts.depth.mean`)) +
  xlab("FRoH (downsampled to 6x, no transitions)") +
  ylab("FRoH (full depth, including transitions)") +
  geom_smooth(method = "lm", color = "black") +
  theme_classic() +
  theme(
    legend.position = "none"
  )

plot_grid(hz_full, f_full, ncol = 2)
ggsave("../figures/suppfig-hetdepth.png", width = 10, height = 5, dpi = 600)
```