---
title: "Population structure"
format:
  html:
    toc: true
---

This notebook contains the code used to make tables and figures relating to
population structure analyses. This includes both main text and supplementary
figures and tables.

```{r}
#| label: load-packages

library(ggplot2) # For general plotting
library(data.table) # For reading in data tables without reading in all columns
library(stringr) # For subsetting metadata strings
library(cowplot) # For arranging plots
library(gridExtra) # For plotting grobs for admix
library(dplyr) # For merging dataframes
library(ggh4x) # For nested ggplot facets
library(ggtext) # For markdown syntax in ggplot
library(ggpubr) #for bgcolor

species <- c("picarus", "pargus", "csemiargus")
genomes <- c("ilPolIcar1.1", "ilPleArgu1.3", "ilCyaSemi1.1")
```

## PCA

Plot the outputs of a principal component analysis for each species together.
This is used for making panel A of Figure 2 and as written makes the lower panel
of Figure S4. To generate the upper panel of Figure S4, I ran the same code
block, but exchanged the path for the covariance matrix to point to the bwa aln
aligned dataset instead.

```{r}
#| label: fig-pca

regs <- c("WSkane", "SWSkane", "ESkane", "SESkane", "NSmaland", "Oland")
regs_named <- c(
  "W Skåne", "SW Skåne", "E Skåne", "SE Skåne", "Småland",
  "Öland"
)
regs_cols <- c("#244F6E", "#244F6E", "#658399", "#658399", "#8A2A3B", "#fecd03")

pcas <- list()

# read in PCAngsd covariance results and compute PCs
for (i in c(1:length(species))) {
  s <- species[i]
  g <- genomes[i]
  cov <- as.matrix(read.table(
    paste0(
      "../analyses/results/datasets/historical-",
      s,
      "-vg-notrans/analyses/pcangsd/historical-",
      s,
      "-vg-notrans.",
      g,
      "_all_excl_pca-admix.dp8_allsites-filts.cov"
    ),
    header = FALSE,
  ))
  pop <- fread(
    paste0(
      "../analyses/results/datasets/historical-",
      s,
      "-vg-notrans/poplists/historical-",
      s,
      "-vg-notrans_all_excl_pca-admix.dp8.indiv.list"
    ),
    header = TRUE,
    sep = "\t",
    select = c("sample", "population", "time")
  )
  e <- eigen(cov)
  pc_var <- round((e$values / sum(e$values)) * 100, 2)
  pc <- data.frame(e$vectors)
  pc$sample <- pop$sample
  pc <- merge(pc, pop, by = "sample")
  pc$species <- s
  pc$region <- str_sub(pc$population, 1, -5)
  species_regs_index <- match(unique(pc$region), regs)
  species_regs <- regs[species_regs_index]
  species_regs_names <- regs_named[species_regs_index]
  species_regs_colors <- regs_cols[species_regs_index]

  pc$species <- factor(
    pc$species,
    levels = species,
    labels = c("Po. icarus", "Pl. argus", "Cy. semiargus")
  )

  pc$time <- factor(
    pc$time,
    levels = c("modern", "historical"),
    labels = c("Modern", "Historical")
  )

  pc$region <- factor(
    pc$region,
    levels = species_regs,
    labels = species_regs_names
  )

# make a ggplot for the PCA of each species
  pcas[[s]] <- ggplot(pc, aes(
    x = X1, y = X2, color = region, shape = time,
    fill = region
  )) +
    facet_grid(cols = vars(species)) +
    geom_point(size = 4, stroke = 1) +
    scale_shape_manual(
      values = c(21, 24)
    ) +
    scale_color_manual(
      values = species_regs_colors,
      name = "Geographic Region"
    ) +
    scale_fill_manual(
      values = alpha(species_regs_colors, 0.7)
    ) +
    xlab(paste0("PC1 (", pc_var[1], "%)")) +
    ylab(paste0("PC2 (", pc_var[2], "%)")) +
    theme_bw() +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      strip.text.x = element_text(size = 10, face = "italic"),
      legend.position = "none",
      aspect.ratio = 1
    )
}

# make combined plot of all three species pcas
pcas <- plot_grid(
  pcas[["picarus"]],
  pcas[["pargus"]],
  pcas[["csemiargus"]],
  ncol = 3, nrow = 1
)
```

# Best fit K admixture plot

Plot the outputs of the best fit K for each species in the admixture analyes. If
no K > 1 converged so K = 1 was selected, plot K = 2 instead as K = 1 is boring
and uninformative.

```{r}
#| label: fig-admixture

# set best fit Ks based on convergence of replicates (Table S8)
best_fit_ks <- c(1, 1, 5)

regs <- c(
  "Skane", "WSkane", "SWSkane", "ESkane", "SESkane", "NSmaland", "Oland"
)
regs_named <- c(
  "Skåne", "W Skåne", "SW Skåne", "E Skåne", "SE Skåne", "Småland", "Öland"
)

# this color list is manually arranged to color clusters in a way that aligns
# with the regional colors in Figure 1. it nees to be rearranged depending on
# the replicate, even converged ones, as the cluster indices is arbitrary
colors <- list()
colors[["picarus"]] <- c("#fecd03", "#244F6E")
colors[["pargus"]] <- c("#244F6E", "#8A2A3B")
colors[["csemiargus"]] <- c("#658399", "#A7B8C5", "#244F6E", "#8A2A3B", "#B5B6B7")


# read in admixture clusters for best fit K of each species and do some metadata
# joining and cleanup for plotting
admix <- list()

for (i in c(1:length(species))) {
  s <- species[i]
  g <- genomes[i]
  k <- best_fit_ks[i]

  if (k == 1) {
    k <- 2
  }

  qopt <- read.table(paste0(
    "../analyses/results/datasets/historical-",
    s,
    "-vg-notrans/analyses/ngsadmix/historical-",
    s,
    "-vg-notrans.",
    g,
    "_all_excl_pca-admix.dp8_allsites-filts_K",
    k,
    ".qopt"
  ))

  pop <- as.data.frame(fread(
    paste0(
      "../analyses/results/datasets/historical-",
      s,
      "-vg-notrans/poplists/historical-",
      s,
      "-vg-notrans_all_excl_pca-admix.dp8.indiv.list"
    ),
    header = TRUE,
    sep = "\t",
    select = c("sample", "population", "time")
  ))

  pop[pop$population == "ESkane1917", "population"] <- "Skane"
  pop[pop$population == "WSkane1943", "population"] <- "Skane"

  qopt$sample <- pop$sample

  qopt <- reshape2::melt(as.data.table(inner_join(qopt, pop)))
  qopt$region <- str_sub(qopt$population, 1, -5)
  qopt$year <- str_sub(qopt$population, -4, -1)
  qopt[qopt$region == "S", "year"] <- "1917/43"
  qopt[qopt$region == "S", "region"] <- "Skane"
  species_regs_index <- match(unique(qopt$region), regs)
  species_regs <- regs[sort(species_regs_index)]
  species_regs_names <- regs_named[sort(species_regs_index)]

  qopt$time <- factor(
    qopt$time,
    levels = c("historical", "modern"),
    labels = c("Historical", "Modern")
  )

  qopt$region <- factor(
    qopt$region,
    levels = species_regs,
    labels = species_regs_names
  )

  qopt$population <- paste0(qopt$region, "\n", qopt$year)

  pops <- unique(qopt$population)

  qopt$population <- factor(
    qopt$population,
    levels = pops[order(match(str_extract(pops, "[^\n]+"), regs_named))]
  )

  # In Pl. argus, make sure that the Skåne samples are in the right order in
  # the plot (i.e. 1917 before 43 to line up with the year label)
  if (s == "pargus") {
    qopt$sample <- factor(
      qopt$sample,
      levels = c("MZLU153003", unique(qopt[qopt$sample != "MZLU153003", ]$sample))
    )
  }

  sp_name <- c("Po. icarus", "Pl. argus", "Cy. semiargus")[i]

  admix[[s]] <- ggplot(qopt, aes(value, fill = variable, x = sample)) +
    geom_bar(position = "fill", stat = "identity", width = 1, color = "black") +
    facet_nested(~ time + population,
      scales = "free", space = "free",
      nest_line = TRUE
    ) +
    ylab(paste0("*", sp_name, "*<br>K = ", k)) +
    guides(fill = "none") +
    scale_y_continuous(expand = c(0, 0), breaks = seq(0, 1, by = 0.2)) +
    scale_fill_manual(values = colors[[s]]) +
    theme_classic() +
    theme(
      axis.title.y = element_markdown(lineheight = 1.2),
      strip.text.x = element_text(size = 8),
      strip.background = element_blank(),
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.line.x = element_blank(),
      axis.ticks.x = element_blank(),
      plot.margin = unit(c(0, 0.1, 0.1, 0.3), "cm")
    )
}

admixes <- plot_grid(
  admix[["picarus"]],
  admix[["pargus"]],
  admix[["csemiargus"]],
  ncol = 1, nrow = 3
)
```

### All K plots

Additionally, generate admix plots for all values of K per species to put in the
supplementary materials. Does the same as above, but loops over the Ks and
arranges them into a stack for Figure S6. At the same time, it generates
correlation of residuals plots using the `plotCorRes` function in the
`visFuns.R` script included in evalAdmix which gets sourced in this block.

```{r}
#| label: suppfig-admix
#| fig-height: 8
#| fig-width: 6

# Read in evaladmix functions
source("https://raw.githubusercontent.com/GenisGE/evalAdmix/2a51aebaca70c9d3b5fb359d6c5c40145c58fce5/visFuns.R")

for (i in c(1:length(species))) {
  s <- species[i]
  g <- genomes[i]

  admix <- list()
  resids <- list()

  sp_name <- c("Po. icarus", "Pl. argus", "Cy. semiargus")[i]

  pop <- as.data.frame(fread(
    paste0(
      "../analyses/results/datasets/historical-",
      s,
      "-vg-notrans/poplists/historical-",
      s,
      "-vg-notrans_all_excl_pca-admix.dp8.indiv.list"
    ),
    header = TRUE,
    sep = "\t",
    select = c("sample", "population", "time")
  ))

  pop[pop$population == "ESkane1917", "population"] <- "Skane"
  pop[pop$population == "WSkane1943", "population"] <- "Skane"

  for (k in c(2:8)) {
    qopt <- read.table(paste0(
      "../analyses/results/datasets/historical-",
      s,
      "-vg-notrans/analyses/ngsadmix/historical-",
      s,
      "-vg-notrans.",
      g,
      "_all_excl_pca-admix.dp8_allsites-filts_K",
      k,
      ".qopt"
    ))

    q <- read.table(
      paste0(
        "../analyses/results/datasets/historical-",
        s,
        "-vg-notrans/analyses/ngsadmix/historical-",
        s,
        "-vg-notrans.",
        g,
        "_all_excl_pca-admix.dp8_allsites-filts_K",
        k,
        ".qopt"
      ),
      stringsAsFactors = TRUE
    )

    qopt$sample <- pop$sample

    qopt <- reshape2::melt(as.data.table(inner_join(qopt, pop)))
    qopt$region <- str_sub(qopt$population, 1, -5)
    qopt$year <- str_sub(qopt$population, -4, -1)
    qopt[qopt$region == "S", "year"] <- "1917/43"
    qopt[qopt$region == "S", "region"] <- "Skane"
    species_regs_index <- match(unique(qopt$region), regs)
    species_regs <- regs[sort(species_regs_index)]
    species_regs_names <- regs_named[sort(species_regs_index)]

    qopt$time <- factor(
      qopt$time,
      levels = c("historical", "modern"),
      labels = c("Historical", "Modern")
    )

    qopt$region <- factor(
      qopt$region,
      levels = species_regs,
      labels = species_regs_names
    )

    qopt$origpop <- qopt$population
    qopt$population <- paste0(qopt$region, "\n", qopt$year)

    pops <- unique(qopt$population)

    qopt$population <- factor(
      qopt$population,
      levels = pops[order(match(str_extract(pops, "[^\n]+"), regs_named))]
    )

    # In Pl. argus, make sure that the Skåne samples are in the right order in
    # the plot (i.e. 1917 before 43 to line up with the year label)
    if (s == "pargus") {
      qopt$sample <- factor(
        qopt$sample,
        levels = c("MZLU153003", unique(qopt[qopt$sample != "MZLU153003", ]$sample))
      )
    }

    if (k == 2) {
      striptext <- element_text(size = 8)
    } else {
      striptext <- element_text(size = 0)
    }

    admix[[k]] <- ggplot(qopt, aes(value, fill = variable, x = sample)) +
      geom_bar(position = "fill", stat = "identity", width = 1, color = "black") +
      facet_nested(~ time + population,
        scales = "free", space = "free",
        nest_line = TRUE
      ) +
      ylab(paste0("K = ", k)) +
      guides(fill = "none") +
      scale_y_continuous(expand = c(0, 0), breaks = seq(0, 1, by = 0.2)) +
      theme_classic() +
      theme(
        axis.title.y = element_markdown(lineheight = 1.2),
        strip.text.x = striptext,
        strip.background = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.line.x = element_blank(),
        axis.ticks.x = element_blank(),
        plot.margin = unit(c(0, 0, 0, 0), "cm")
      )

    ord <- orderInds(
      pop = as.vector(qopt[qopt$variable == "V1", ]$population),
      popord = pops[order(match(str_extract(pops, "[^\n]+"), regs_named))]
    )

    r <- as.matrix(
      read.table(
        paste0(
          "../analyses/results/datasets/historical-",
          s,
          "-vg-notrans/analyses/ngsadmix/historical-",
          s,
          "-vg-notrans.",
          g,
          "_all_excl_pca-admix.dp8_allsites-filts_K",
          k,
          ".corres"
        )
      )
    )

    png(
      paste0("../figures/suppfig-evaladmix-", s, "-K", k, ".png"),
      height = 600,
      width = 700
    )
    plotCorRes(
      cor_mat = r,
      pop = as.vector(qopt[qopt$variable == "V1", ]$population),
      ord = ord,
      title = paste0("K=", k),
      rotatelabpop = 45,
      min_z = -0.1,
      max_z = 0.1
    )
    dev.off()
  }
  title <- ggdraw() +
    draw_label(
      sp_name,
      fontface = "bold.italic",
      x = 0,
      hjust = 0
    ) +
    theme(
      plot.margin = margin(0, 0, 0, 7)
    )
  plot_grid(
    title,
    admix[[2]],
    admix[[3]],
    admix[[4]],
    admix[[5]],
    admix[[6]],
    admix[[7]],
    admix[[8]],
    nrow = 8,
    ncol = 1,
    rel_heights = c(0.2, 1, rep(0.6, 8))
  ) + ggtitle(sp_name)
  ggsave(paste0(
    "../figures/suppfig-admix-", s, ".png"
  ), dpi = 300, create.dir = TRUE)
}
```

## Combined PCA + Admixture plot (Figure 2)

```{r}
#| label: fig-structure
#| fig-width: 10
#| fig-height: 9

plot_grid(
  pcas,
  admixes,
  ncol = 1,
  nrow = 2,
  rel_heights = c(2, 2.5),
  labels = c("AUTO"),
  scale = c(1, 0.99)
) + bgcolor("white")

ggsave("../figures/figure2.png", dpi = 600, create.dir = TRUE)
ggsave("../figures/figure2.svg", dpi = 600, create.dir = TRUE)
```