---
title: "Putatively deleterious mutation burden"
format:
  html: default
---

This notebook contains the code used to make tables and figures relating to
putative deleterious mutation burden.

```{r}
#| label: load-packages

set.seed(25)

library(ggplot2) # For general plotting
library(data.table) # For reading in data tables without reading in all columns
library(dplyr) # For manipulating dataframes
library(tidyr) # For splitting data frame columns
library(stringr) # For subsetting metadata strings
library(ggtext) # For markdown syntax in ggplot
library(ggpubr) # For t.test results within plots
library(scales) # For muted colors
library(ggh4x) # For nested ggplot facets
library(cowplot) # For arranging plots
library(broom.mixed)
library(nlme)
library(bootstrap)

species <- c("picarus", "pargus", "csemiargus")
species_names <- c("Po. icarus", "Pl. argus", "Cy. semiargus")
genomes <- c("ilPolIcar1.1", "ilPleArgu1.3", "ilCyaSemi1.1")
```

## Putatively deleterious mutation burden

### Derived count statistics and table

In this study as a genomic proxy for genetic load, we used the derived count of
putatively deleterious mutations. The main workflow provides this output as a
table of derived allele counts at different position categories, including
counts in homozygous state in particular. We first compile these counts into
derived counts / 1000 genotyped positions to make them comparable across samples
with variable missingness. We calculate means of these counts for each mutation
category both for the total count and homozygous count. Then we performa linear
mixed models with derived count as a response to sampling time period, with
sampling region as a random factor. We also get confidence intervals around the
mean counts with jackknifing.

```{r}
#| label: stats-burden

# read in the derived count tables as output by the workflow, using the one
# that is polarized by the inferred ancestral states. read in Froh estimates
# as well to associate them later
varcounts_all <- c()
froh_all <- c()

for (i in c(1:length(species))) {
  s <- species[i]
  g <- genomes[i]

  # Read in variant impact counts
  varcounts <- read.table(
    paste0(
      "../analyses/results/datasets/historical-",
      s,
      "-vg-notrans/analyses/burden/historical-",
      s,
      "-vg-notrans.",
      g,
      "_all.dp6_allsites-filts.filtered_mindp5-biallelic-jointcall_allbal0.21-0.79.notrans.fmiss0.3.varimpacts_anc.tsv"
    ),
    header = TRUE
  )

  froh <- fread(
    paste0(
      "../analyses/results/datasets/historical-",
      s,
      "-vg-notrans/plots/inbreeding/historical-",
      s,
      "-vg-notrans.",
      g,
      "_all.dp6_allsites-filts.filtered_mindp5-biallelic-jointcall_allbal0.21-0.79.notrans.fmiss0.3.bcftools.ind_froh.tsv"
    ),
    header = TRUE,
    sep = "\t",
    select = c("sample", "population", "time", "depth", "Froh")
  )
  varcounts <- merge(varcounts, froh, by = c("sample", "population", "time", "depth"))
  varcounts$species <- s
  froh$species <- s
  froh_all <- rbind(froh_all, froh)
  varcounts_all <- rbind(varcounts_all, varcounts)
}

varcounts_all$year <- str_sub(varcounts_all$population, -4, -1)
varcounts_all$region <- str_sub(varcounts_all$population, 1, -5)

varcounts_all$species <- factor(
  varcounts_all$species,
  levels = species,
  labels = c("Po. icarus", "Pl. argus", "Cy. semiargus")
)

varcounts_all$time <- factor(
  varcounts_all$time,
  levels = c("historical", "modern"),
  labels = c("Historical", "Modern")
)

varcounts_all$region <- factor(
  varcounts_all$region,
  levels = c("WSkane", "SWSkane", "ESkane", "SESkane", "NSmaland", "Oland"),
  labels = c(
    "W Skåne", "SW Skåne", "E Skåne", "SE Skåne", "Småland", "Öland"
  )
)

varcounts_all$region_group <- varcounts_all$region
varcounts_all[varcounts_all$region_group == "SW Skåne", ]$region_group <- "W Skåne"
varcounts_all[varcounts_all$region_group == "SE Skåne", ]$region_group <- "E Skåne"

# perform linear models for derived count as a response to time period across
# relevant categories of mutations. these include putatively deleterious
# categories (high impact, moderate impact, conserved) as well as putatively
# neutral categories (low impact, intergenic) as a control. always normalize
# by total genotype count (nsites)
varcounts_stats <- c()

for (s in species_names) {
  for (i in c("gerp_n", "high", "mod", "low", "inter", "gerp_hom_n", "high_hom", "mod_hom", "low_hom", "inter_hom")) {
    hist <- mean(varcounts_all[varcounts_all$species == s & varcounts_all$time == "Historical", i] * 1000 / varcounts_all[varcounts_all$species == s & varcounts_all$time == "Historical", "nsites"])
    mod <- mean(varcounts_all[varcounts_all$species == s & varcounts_all$time == "Modern", i] * 1000 / varcounts_all[varcounts_all$species == s & varcounts_all$time == "Modern", "nsites"])
    ratio <- mod / hist
    model <- lme(as.formula(paste0(i, " * 1000 / nsites ~ time")),
      random = ~ 1 | region_group,
      data = varcounts_all[varcounts_all$species == s, ]
    )
    tidyed <- tidy(model, conf.int = TRUE, effects = "fixed")
    tidydf <- tidyed[2, c(3, 4, 6, 7, 8, 9)]
    tidydf$hist.mean <- hist
    jackknife <- quantile(jackknife(varcounts_all[varcounts_all$species == s & varcounts_all$time == "Historical", i] * 1000 / varcounts_all[varcounts_all$species == s & varcounts_all$time == "Historical", "nsites"], mean)$jack.values, c(0.025, 0.975))
    tidydf$hist.low.jk.ci <- jackknife[1]
    tidydf$hist.high.jk.ci <- jackknife[2]
    tidydf$mod.mean <- mod
    jackknife <- quantile(jackknife(varcounts_all[varcounts_all$species == s & varcounts_all$time == "Modern", i] * 1000 / varcounts_all[varcounts_all$species == s & varcounts_all$time == "Modern", "nsites"], mean)$jack.values, c(0.025, 0.975))
    tidydf$mod.low.jk.ci <- jackknife[1]
    tidydf$mod.high.jk.ci <- jackknife[2]
    tidydf$`mod/hist` <- ratio
    tidydf <- round(tidydf, 5)
    tidydf$species <- s
    tidydf <- tidydf[, c(14, 7, 8, 9, 10, 11, 12, 13, 1, 2, 3, 5, 6, 4)]
    tidydf$type <- i
    varcounts_stats <- rbind(varcounts_stats, tidydf)
  }
}

# write out Table S11
write.table(varcounts_stats, "../tables/burden_stats.tsv",
  row.names = FALSE, quote = FALSE, sep = "\t"
)
```

### Derived count main boxplots

The next block plots boxplots for each mutation category comparing historical
derived counts with moderns. A separate panel is made for each species, and
plots are grouped by total and homozygous derived count.

```{r}
#| label: fig-burden

high <- ggplot(varcounts_all, aes(x = time, y = high * 1000 / nsites)) +
  facet_grid(cols = vars(species)) +
  geom_boxplot(outliers = FALSE) +
  geom_point(
    aes(color = Froh, group = region),
    position = position_jitterdodge(jitter.width = 0.05)
  ) +
  scale_color_gradient(high = "#6e016b", low = "#9ebcda") +
  ylab("LoF") +
  scale_y_continuous(
    labels = number_format(accuracy = 0.01),
    expand = expansion(mult = c(0.05, 0.20))
  ) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title.x = element_blank(),
    strip.text.x = element_text(size = 10, face = "italic")
  )

high_hom <- ggplot(varcounts_all, aes(x = time, y = high_hom * 1000 / nsites)) +
  facet_grid(cols = vars(species)) +
  geom_boxplot(outliers = FALSE) +
  geom_point(
    aes(color = Froh, group = region),
    position = position_jitterdodge(jitter.width = 0.05)
  ) +
  scale_color_gradient(high = "#6e016b", low = "#9ebcda") +
  scale_y_continuous(
    labels = number_format(accuracy = 0.01),
    expand = expansion(mult = c(0.05, 0.20))
  ) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title.y = element_blank(),
    axis.title.x = element_blank(),
    strip.text.x = element_text(size = 10, face = "italic")
  )

mod <- ggplot(varcounts_all, aes(x = time, y = mod * 1000 / nsites)) +
  facet_grid(cols = vars(species)) +
  geom_boxplot(outliers = FALSE) +
  geom_point(
    aes(color = Froh, group = region),
    position = position_jitterdodge(jitter.width = 0.05)
  ) +
  scale_color_gradient(high = "#6e016b", low = "#9ebcda") +
  ylab("Missense") +
  scale_y_continuous(
    labels = number_format(accuracy = 0.1),
    expand = expansion(mult = c(0.05, 0.20))
  ) +
  geom_signif(
    data = varcounts_all[varcounts_all$species == "Pl. argus", ],
    comparisons = list(c("Historical", "Modern")),
    annotations = "***"
  ) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title.x = element_blank(),
    strip.text.x = element_text(size = 10, face = "italic")
  )

mod_hom <- ggplot(varcounts_all, aes(x = time, y = mod_hom * 1000 / nsites)) +
  facet_grid(cols = vars(species)) +
  geom_boxplot(outliers = FALSE) +
  geom_point(
    aes(color = Froh, group = region),
    position = position_jitterdodge(jitter.width = 0.05)
  ) +
  scale_color_gradient(high = "#6e016b", low = "#9ebcda") +
  scale_y_continuous(
    labels = number_format(accuracy = 0.1),
    expand = expansion(mult = c(0.05, 0.20))
  ) +
  geom_signif(
    data = varcounts_all[varcounts_all$species == "Cy. semiargus", ],
    comparisons = list(c("Historical", "Modern")),
    annotations = "**"
  ) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title.y = element_blank(),
    axis.title.x = element_blank(),
    strip.text.x = element_text(size = 10, face = "italic")
  )


gerp <- ggplot(varcounts_all, aes(x = time, y = gerp_n * 1000 / nsites)) +
  facet_grid(cols = vars(species)) +
  geom_boxplot(outliers = FALSE) +
  geom_point(
    aes(color = Froh, group = region),
    position = position_jitterdodge(jitter.width = 0.05)
  ) +
  scale_color_gradient(high = "#6e016b", low = "#9ebcda") +
  ylab("Conserved Positions") +
  scale_y_continuous(
    labels = number_format(accuracy = 0.1),
    expand = expansion(mult = c(0.05, 0.20))
  ) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title.x = element_blank(),
    strip.text.x = element_text(size = 10, face = "italic")
  )

gerp_hom <- ggplot(varcounts_all, aes(x = time, y = gerp_hom_n * 1000 / nsites)) +
  facet_grid(cols = vars(species)) +
  geom_boxplot(outliers = FALSE) +
  geom_point(
    aes(color = Froh, group = region),
    position = position_jitterdodge(jitter.width = 0.05)
  ) +
  scale_color_gradient(high = "#6e016b", low = "#9ebcda") +
  scale_y_continuous(
    labels = number_format(accuracy = 0.1),
    expand = expansion(mult = c(0.05, 0.20))
  ) +
  geom_signif(
    data = varcounts_all[varcounts_all$species %in% c("Pl. argus", "Cy. semiargus"), ],
    comparisons = list(c("Historical", "Modern")),
    annotations = "**"
  ) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title.x = element_blank(),
    strip.text.x = element_text(size = 10, face = "italic")
  )
```

## Inbreeding coefficients partitiond by RoH size

To visualize how much of the inbreeding coefficient was composed by runs of
homozygosity by different lengths, we plotted them as bars colored by the
proportion coming from different size classes.

```{r}
#| label: fig-csemiargus-roh

rohs <- read.table("../analyses/results/datasets/historical-csemiargus-vg-notrans/analyses/roh/bcftools/historical-csemiargus-vg-notrans.ilCyaSemi1.1_all.dp6_allsites-filts.filtered_mindp5-biallelic-jointcall_allbal0.21-0.79.notrans.fmiss0.3.regs.roh")

colnames(rohs) <- c(
  "type", "sample", "chr", "start", "end", "length",
  "inform_sites", "phred"
)

# exclude ROH < 200kb and < 30 Pherd score
rohs <- rohs[rohs$length >= 200000 & rohs$phred >= 30, ]

nroh <- rohs %>%
  group_by(sample) %>%
  summarise(cumroh = sum(length), nroh = n())

inds <- read.table(
  "../analyses/results/datasets/historical-csemiargus-vg-notrans/poplists/historical-csemiargus-vg-notrans_all.indiv.list",
  header = TRUE
)

# drop uncalled sample from ind list for sample looping
inds <- inds[inds$sample != "MZLU152765", ]

nroh <- merge(inds, nroh, by = "sample", all = TRUE)
nroh[is.na(nroh)] <- 0

norun <- c()
norun$type <- rep("RG", nrow(inds))
norun$sample <- inds$sample
norun$chr <- rep(0, nrow(inds))
norun$start <- rep(0, nrow(inds))
norun$end <- rep(0, nrow(inds))
norun$length <- rep(0, nrow(inds))
norun$inform_sites <- rep(0, nrow(inds))
norun$phred <- rep(0, nrow(inds))
norun <- data.frame(norun)

# set total autosomal length for C. semiargus
autosize <- 416794061

# prepare size bins to loop over
bins <- c(Inf, 6000000, 3000000, 1500000, 500000, 200000)

frohs <- c()
ranges <- c()

# loop over bins, quantifying Froh for that size bin in each sample
for (i in seq(length(bins))) {
  if (i <= length(bins) - 1) {
    runs <- rohs[rohs$length < bins[i], ]
    runs <- runs[runs$length >= bins[i + 1], ]
    runs <- rbind(runs, norun)
    runs <- runs %>%
      group_by(sample) %>%
      summarize(length = sum(length))
    runs <- merge(runs, inds, by = "sample")
    runs$froh <- runs$length / autosize
    runs$range <- paste0(bins[i], "-", bins[i + 1])
    ranges <- c(ranges, paste0(bins[i], "-", bins[i + 1]))
    frohs <- rbind(frohs, runs)
  }
}

range_labels <- rev(c("0.2-0.5", "0.5-1.5", "1.5-3", "3-6", ">6"))
frohs$range <- as.factor(frohs$range)
frohs$range <- factor(frohs$range, levels = ranges, labels = range_labels)

indfroh <- frohs %>%
  group_by(sample) %>%
  summarize(froh = sum(froh))

frohs$sample <- factor(frohs$sample, levels = indfroh$sample[order(indfroh$froh)])

popord <- c(
  "WSkane1951", "ESkane1956", "NSmaland1936", "WSkane2021", "ESkane2021",
  "SESkane2022", "NSmaland2021"
)

poplab <- c(
  "W Skåne\n1951", "E Skåne\n1956", "Småland\n1936", "W Skåne\n2021",
  "E Skåne\n2021", "SE Skåne\n2022", "Småland\n2021"
)

frohs$population <- factor(frohs$population, levels = popord, labels = poplab)
frohs$time <- factor(frohs$time,
  levels = c("historical", "modern"),
  labels = c("Historical", "Modern")
)

frohplot <- ggplot(
  data = frohs, aes(x = sample, y = froh, alpha = range), fill = "black"
) +
  geom_bar(position = "stack", stat = "identity") +
  facet_nested(~ time + population,
    scales = "free_x", space = "free",
    nest_line = TRUE
  ) +
  theme_bw() +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 0.35)) +
  scale_alpha_discrete(range = c(1, 0.4)) +
  labs(y = "*F~RoH~* (*Cy. semiargus*)", alpha = "*RoH* size (Mb)") +
  theme(
    legend.position = "bottom",
    strip.background = element_blank(),
    axis.text.x = element_blank(),
    axis.title.y = element_markdown(),
    legend.title = element_markdown(),
    axis.title.x = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
  )
```

## Figure 3

Combine Froh barplots and derived count boxplots into Figure 3

```{r}
#| label: fig-load
#| fig-height: 8
#| fig-width: 8

# adjust plots so that they drop info that is depicted primarily by another
# panel as well as fix font sizes and borders for paneling
mod_clean <- mod +
  ggtitle("Total derived count") +
  theme(
    plot.margin = unit(c(5.5, 5.5, 3, 5.5), "pt"),
    legend.position = "none",
    axis.title.y = element_markdown(size = 8),
    axis.text.x = element_blank(),
    strip.background = element_blank()
  )

high_clean <- high +
  theme(
    plot.margin = unit(c(3, 5.5, 3, 5.5), "pt"),
    legend.position = "none",
    strip = element_blank(),
    axis.title.y = element_markdown(size = 8),
    axis.text.x = element_blank(),
    strip.text.x = element_blank(),
    strip.background = element_blank()
  )

gerp_clean <- gerp +
  theme(
    plot.margin = unit(c(3, 5.5, 5.5, 5.5), "pt"),
    legend.position = "none",
    strip = element_blank(),
    axis.title.y = element_markdown(size = 8),
    strip.text.x = element_blank(),
    strip.background = element_blank()
  )

mod_hom_clean <- mod_hom +
  ggtitle("Homozygous derived count") +
  theme(
    plot.margin = unit(c(5.5, 5.5, 3, 5.5), "pt"),
    legend.position = "none",
    axis.title.y = element_blank(),
    axis.text.x = element_blank(),
    strip.background = element_blank()
  )

high_hom_clean <- high_hom +
  theme(
    plot.margin = unit(c(3, 5.5, 3, 5.5), "pt"),
    legend.position = "none",
    strip = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_blank(),
    strip.text.x = element_blank(),
    strip.background = element_blank()
  )

gerp_hom_clean <- gerp_hom +
  theme(
    plot.margin = unit(c(3, 5.5, 5.5, 5.5), "pt"),
    legend.position = "none",
    strip = element_blank(),
    axis.title.y = element_blank(),
    strip.text.x = element_blank(),
    strip.background = element_blank()
  )

total_load <- plot_grid(
  mod_clean,
  high_clean,
  gerp_clean,
  rel_heights = c(1.2, 0.9, 1),
  ncol = 1,
  align = "v"
)

real_load <- plot_grid(
  mod_hom_clean,
  high_hom_clean,
  gerp_hom_clean,
  rel_heights = c(1.2, 0.9, 1),
  ncol = 1,
  align = "v"
)

load <- plot_grid(total_load, real_load, ncol = 2, labels = c("B", "C"), rel_widths = c(1.1, 1))

legend_inb <- get_plot_component(
  gerp +
    labs(color = "*F~RoH~*") +
    theme(
      legend.position = "bottom",
      legend.title = element_markdown(),
      legend.key.width = unit(2, "cm")
    ),
  "guide-box",
  return_all = TRUE
)[[3]]

plot_grid(frohplot, load, legend_inb,
  ncol = 1,
  rel_heights = c(0.4, 0.7, 0.1),
  labels = c("A", "", "")
) + bgcolor("white")

ggsave("../figures/figure3.png", dpi = 600, create.dir = TRUE)
ggsave("../figures/figure3.svg", dpi = 600, create.dir = TRUE)
```

## Associate inbreeding coefficient with derived count

As the increased homozygosity of weakly deleterious mutations is expected with
inbreeding, we assessed the relationship of the derived count with individual
inbreeding coefficients in the modern *Cy. semiargus* samples. The below code
plots this association for the downsampled transitions removed data and also the
full depth data with and without transitions.

```{r}
#| label: fig-inbrload
#| fig-width: 8
#| fig-height: 9

# save the downsampled derived counts into a new object with just csemiargus
var_csemi <- varcounts_all[varcounts_all$species == "Cy. semiargus",]

# plot regressions of the various putatively deleterious categories for the 6x
# data with the derived count on y and inbreeding coefficient on x
# use total count as default, colored red
gerpreg_6x <- ggplot(var_csemi, aes(x = Froh, y = gerp_n * 1000 / nsites)) +
  ggtitle("Conserved Regions") +
  geom_point(color = "red") +
  # add homozygous count in blue, do the same for linear models and ther metrics
  geom_point(aes(y = gerp_hom_n * 1000 / nsites), color = "blue") +
  geom_smooth(method = "lm", color = "red") +
  geom_smooth(aes(y = gerp_hom_n * 1000 / nsites), method = "lm", color = "blue") +
  ylab("Derived per 1000 GTs") +
  xlab("Inbreeding Coefficient (*F~RoH~*)") +
  ylim(c(0,2.5)) +
  stat_cor(color = "red", label.y = 1.6) +
  stat_regline_equation(color = "red", label.y = 1.4) +
  stat_cor(aes(y = gerp_hom_n * 1000 / nsites), color = "blue", label.y = 1, label.x = 0.11) +
  stat_regline_equation(aes(y = gerp_hom_n * 1000 / nsites), color = "blue", label.y = 0.8, label.x = 0.11) +
  theme_classic() +
  theme(
    axis.title.x = element_markdown(),
    legend.position = "none"
  )


highreg_6x <- ggplot(var_csemi, aes(x = Froh, y = high * 1000 / nsites)) +
  ggtitle("LoF") +
  geom_point(color = "red") +
  geom_point(aes(y = high_hom * 1000 / nsites), color = "blue") +
  geom_smooth(method = "lm", color = "red") +
  geom_smooth(aes(y = high_hom * 1000 / nsites), method = "lm", color = "blue") +
  ylab("Derived per 1000 GTs") +
  xlab("Inbreeding Coefficient (*F~RoH~*)") +
  ylim(0, 0.15) +
  stat_cor(color = "red", label.y = 0.14) +
  stat_regline_equation(color = "red", label.y = 0.12) +
  stat_cor(aes(y = high_hom * 1000 / nsites), label.y = 0.01, label.x = 0.11, color = "blue") +
  stat_regline_equation(aes(y = high_hom * 1000 / nsites), color = "blue", label.y = 0, label.x = 0.11) +
  theme_classic() +
  theme(
    axis.title.x = element_markdown(),
    legend.position = "none"
  )

modreg_6x <- ggplot(var_csemi, aes(x = Froh, y = mod * 1000 / nsites)) +
  ggtitle("Missense") +
  geom_point(color = "red") +
  geom_point(aes(y = mod_hom * 1000 / nsites), color = "blue") +
  geom_smooth(method = "lm", color = "red") +
  geom_smooth(aes(y = mod_hom * 1000 / nsites), method = "lm", color = "blue") +
  ylab("Derived per 1000 GTs") +
  xlab("Inbreeding Coefficient (*F~RoH~*)") +
  ylim(0, 7) +
  stat_cor(color = "red", label.y = 5) +
  stat_regline_equation(color = "red", label.y = 4.5) +
  stat_cor(aes(y = mod_hom * 1000 / nsites), label.y = 0.5, label.x = 0.11, color = "blue") +
  stat_regline_equation(aes(y = mod_hom * 1000 / nsites), color = "blue", label.y = 0, label.x = 0.11) +
  theme_classic() +
  theme(
    axis.title.x = element_markdown(),
    legend.position = "none"
  )

# redo the above supp figure at full depth with transitions, first making a
# new df for the full depth burden counts and froh

s <- "csemiargus"
g <- "ilCyaSemi1.1"

varcounts <- read.table(
  paste0(
    "../analyses/results/datasets/historical-",
    s,
    "-vg-trans-fulldp-modonly/analyses/burden/historical-",
    s,
    "-vg-trans-fulldp-modonly.",
    g,
    "_all_allsites-filts.filtered_mindp5-biallelic-jointcall_allbal0.21-0.79.trans.fmiss0.3.varimpacts_anc.tsv"
  ),
  header = TRUE
)

froh <- fread(
  "../analyses/results/datasets/historical-csemiargus-vg-trans-fulldp-modonly/plots/inbreeding/historical-csemiargus-vg-trans-fulldp-modonly.ilCyaSemi1.1_all_allsites-filts.filtered_mindp5-biallelic-jointcall_allbal0.21-0.79.trans.fmiss0.3.bcftools.ind_froh.tsv",
  header = TRUE,
  sep = "\t",
  select = c("sample", "Froh")
)
varcounts <- merge(varcounts, froh, by = c("sample"))
varcounts$species <- s
varcounts_all <- varcounts

varcounts_all$year <- str_sub(varcounts_all$population, -4, -1)
varcounts_all$region <- str_sub(varcounts_all$population, 1, -5)

varcounts_all$species <- factor(
  varcounts_all$species,
  levels = species,
  labels = c("Po. icarus", "Pl. argus", "Cy. semiargus")
)

varcounts_all$time <- factor(
  varcounts_all$time,
  levels = c("historical", "modern"),
  labels = c("Historical", "Modern")
)

varcounts_all$region <- factor(
  varcounts_all$region,
  levels = c("WSkane", "SWSkane", "ESkane", "SESkane", "NSmaland", "Oland"),
  labels = c(
    "W Skåne", "SW Skåne", "E Skåne", "SE Skåne", "Småland", "Öland"
  )
)

var_csemi <- varcounts_all

gerpreg_fulltrans <- ggplot(var_csemi, aes(x = Froh, y = gerp_n * 1000 / nsites)) +
  ggtitle("Conserved Regions") +
  geom_point(color = "red") +
  geom_point(aes(y = gerp_hom_n * 1000 / nsites), color = "blue") +
  geom_smooth(method = "lm", color = "red") +
  geom_smooth(aes(y = gerp_hom_n * 1000 / nsites), method = "lm", color = "blue") +
  ylab("Derived per 1000 GTs") +
  xlab("Inbreeding Coefficient (*F~RoH~*)") +
  ylim(c(0,2.5)) +
  stat_cor(color = "red", label.y = 1.6) +
  stat_regline_equation(color = "red", label.y = 1.4) +
  stat_cor(aes(y = gerp_hom_n * 1000 / nsites), color = "blue", label.y = 1, label.x = 0.11) +
  stat_regline_equation(aes(y = gerp_hom_n * 1000 / nsites), color = "blue", label.y = 0.8, label.x = 0.11) +
  theme_classic() +
  theme(
    axis.title.x = element_markdown(),
    legend.position = "none"
  )


highreg_fulltrans <- ggplot(var_csemi, aes(x = Froh, y = high * 1000 / nsites)) +
  ggtitle("LoF") +
  geom_point(color = "red") +
  geom_point(aes(y = high_hom * 1000 / nsites), color = "blue") +
  geom_smooth(method = "lm", color = "red") +
  geom_smooth(aes(y = high_hom * 1000 / nsites), method = "lm", color = "blue") +
  ylab("Derived per 1000 GTs") +
  xlab("Inbreeding Coefficient (*F~RoH~*)") +
  ylim(0, 0.15) +
  stat_cor(color = "red", label.y = 0.14) +
  stat_regline_equation(color = "red", label.y = 0.12) +
  stat_cor(aes(y = high_hom * 1000 / nsites), label.y = 0.01, label.x = 0.11, color = "blue") +
  stat_regline_equation(aes(y = high_hom * 1000 / nsites), color = "blue", label.y = 0, label.x = 0.11) +
  theme_classic() +
  theme(
    axis.title.x = element_markdown(),
    legend.position = "none"
  )

modreg_fulltrans <- ggplot(var_csemi, aes(x = Froh, y = mod * 1000 / nsites)) +
  ggtitle("Missense") +
  geom_point(color = "red") +
  geom_point(aes(y = mod_hom * 1000 / nsites), color = "blue") +
  geom_smooth(method = "lm", color = "red") +
  geom_smooth(aes(y = mod_hom * 1000 / nsites), method = "lm", color = "blue") +
  ylab("Derived per 1000 GTs") +
  xlab("Inbreeding Coefficient (*F~RoH~*)") +
  ylim(0, 7) +
  stat_cor(color = "red", label.y = 6) +
  stat_regline_equation(color = "red", label.y = 5.5) +
  stat_cor(aes(y = mod_hom * 1000 / nsites), label.y = 0.5, label.x = 0.11, color = "blue") +
  stat_regline_equation(aes(y = mod_hom * 1000 / nsites), color = "blue", label.y = 0, label.x = 0.11) +
  theme_classic() +
  theme(
    axis.title.x = element_markdown(),
    legend.position = "none"
  )

# redo the above supp figure at full depth without transitions, first making a
# new df for the full depth burden counts and froh

varcounts <- read.table(
  paste0(
    "../analyses/results/datasets/historical-",
    s,
    "-vg-notrans-fulldp-modonly/analyses/burden/historical-",
    s,
    "-vg-notrans-fulldp-modonly.",
    g,
    "_all_allsites-filts.filtered_mindp5-biallelic-jointcall_allbal0.21-0.79.notrans.fmiss0.3.varimpacts_anc.tsv"
  ),
  header = TRUE
)

froh <- fread(
  "../analyses/results/datasets/historical-csemiargus-vg-notrans-fulldp-modonly/plots/inbreeding/historical-csemiargus-vg-notrans-fulldp-modonly.ilCyaSemi1.1_all_allsites-filts.filtered_mindp5-biallelic-jointcall_allbal0.21-0.79.notrans.fmiss0.3.bcftools.ind_froh.tsv",
  header = TRUE,
  sep = "\t",
  select = c("sample", "Froh")
)
varcounts <- merge(varcounts, froh, by = c("sample"))
varcounts$species <- s
varcounts_all <- varcounts

varcounts_all$year <- str_sub(varcounts_all$population, -4, -1)
varcounts_all$region <- str_sub(varcounts_all$population, 1, -5)

varcounts_all$species <- factor(
  varcounts_all$species,
  levels = species,
  labels = c("Po. icarus", "Pl. argus", "Cy. semiargus")
)

varcounts_all$time <- factor(
  varcounts_all$time,
  levels = c("historical", "modern"),
  labels = c("Historical", "Modern")
)

varcounts_all$region <- factor(
  varcounts_all$region,
  levels = c("WSkane", "SWSkane", "ESkane", "SESkane", "NSmaland", "Oland"),
  labels = c(
    "W Skåne", "SW Skåne", "E Skåne", "SE Skåne", "Småland", "Öland"
  )
)

var_csemi <- varcounts_all

gerpreg_fullnotrans <- ggplot(var_csemi, aes(x = Froh, y = gerp_n * 1000 / nsites)) +
  ggtitle("Conserved Regions") +
  geom_point(color = "red") +
  geom_point(aes(y = gerp_hom_n * 1000 / nsites), color = "blue") +
  geom_smooth(method = "lm", color = "red") +
  geom_smooth(aes(y = gerp_hom_n * 1000 / nsites), method = "lm", color = "blue") +
  ylab("Derived per 1000 GTs") +
  xlab("Inbreeding Coefficient (*F~RoH~*)") +
  ylim(c(0,2.5)) +
  stat_cor(color = "red", label.y = 1.6) +
  stat_regline_equation(color = "red", label.y = 1.4) +
  stat_cor(aes(y = gerp_hom_n * 1000 / nsites), color = "blue", label.y = 1, label.x = 0.11) +
  stat_regline_equation(aes(y = gerp_hom_n * 1000 / nsites), color = "blue", label.y = 0.8, label.x = 0.11) +
  theme_classic() +
  theme(
    axis.title.x = element_markdown(),
    legend.position = "none"
  )


highreg_fullnotrans <- ggplot(var_csemi, aes(x = Froh, y = high * 1000 / nsites)) +
  ggtitle("LoF") +
  geom_point(color = "red") +
  geom_point(aes(y = high_hom * 1000 / nsites), color = "blue") +
  geom_smooth(method = "lm", color = "red") +
  geom_smooth(aes(y = high_hom * 1000 / nsites), method = "lm", color = "blue") +
  ylab("Derived per 1000 GTs") +
  xlab("Inbreeding Coefficient (*F~RoH~*)") +
  ylim(0, 0.15) +
  stat_cor(color = "red", label.y = 0.14) +
  stat_regline_equation(color = "red", label.y = 0.12) +
  stat_cor(aes(y = high_hom * 1000 / nsites), label.y = 0.01, label.x = 0.11, color = "blue") +
  stat_regline_equation(aes(y = high_hom * 1000 / nsites), color = "blue", label.y = 0, label.x = 0.11) +
  theme_classic() +
  theme(
    axis.title.x = element_markdown(),
    legend.position = "none"
  )

modreg_fullnotrans <- ggplot(var_csemi, aes(x = Froh, y = mod * 1000 / nsites)) +
  ggtitle("Missense") +
  geom_point(color = "red") +
  geom_point(aes(y = mod_hom * 1000 / nsites), color = "blue") +
  geom_smooth(method = "lm", color = "red") +
  geom_smooth(aes(y = mod_hom * 1000 / nsites), method = "lm", color = "blue") +
  ylab("Derived per 1000 GTs") +
  xlab("Inbreeding Coefficient (*F~RoH~*)") +
  ylim(0, 7) +
  stat_cor(color = "red", label.y = 6) +
  stat_regline_equation(color = "red", label.y = 5.5) +
  stat_cor(aes(y = mod_hom * 1000 / nsites), label.y = 0.5, label.x = 0.11, color = "blue") +
  stat_regline_equation(aes(y = mod_hom * 1000 / nsites), color = "blue", label.y = 0, label.x = 0.11) +
  theme_classic() +
  theme(
    axis.title.x = element_markdown(),
    legend.position = "none"
  )

# plot Figure S10
plot_grid(
  gerpreg_6x, highreg_6x, modreg_6x,
  gerpreg_fullnotrans, highreg_fullnotrans, modreg_fullnotrans,
  gerpreg_fulltrans, highreg_fulltrans, modreg_fulltrans,
  ncol = 3, nrow = 3
)
ggsave("../figures/suppfig-inbreg.png", dpi = 600, create.dir = TRUE)
```
