---
title: "Supplemental Polyommatini species"
format:
  html:
    code-fold: true
---

```{r}
#| label: load-packages

library(data.table)
library(stringr)
library(ggplot2)
library(dplyr)
library(ggpubr) #for bgcolor

main_species_f <- c("picarus-vg-notrans", "pargus-vg-notrans", "csemiargus-vg-notrans")
main_species <- c("picarus", "pargus", "csemiargus")
main_species_m <- c("0.3", "0.3", "0.3")
main_genomes <- c("ilPolIcar1.1", "ilPleArgu1.3", "ilCyaSemi1.1")

species_f <- c(
  "aagestis-vg", "aoptilete-vg", #"palcon",
  "cminimus-vg", "pargyrognomon-vg", "sorion-vg"
)

species <- c(
  "aagestis", "aoptilete", #"palcon",
  "cminimus", "pargyrognomon", "sorion"
)
genomes <- c(
  "ilAriAges2.1", "ilAgrOpti3.hap1.1", # "ilCelArgi3.2",
  "ilCupMini1.1", "ilPleArgu1.3-pargyrognomon", "ilScoOrio1.hap1.1"
)
species_m <- c(
  "0.4", "0.4", "0.4", "0.4", "0.4"
)

```

## Heterozygosity change per year

One of the metrics suggested for utilization in Sweden is the per year change in
heterozygosity. While the warning categories are suggested only for significant
differences, which we can't really check with single samples per time period, we
use this same calculation as it is a useful visualization.

```{r}
#| label: main-species-het

# As the main species have specific considerations (i.e. multiple sampling
# areas), they'll be processed separately first.

relhet_all <- c()

  for (i in seq_along(c(main_species, species))) {
    s <- c(main_species, species)[i]
    f <- c(main_species_f, species_f)[i]
    g <- c(main_genomes, genomes)[i]
    m <- c(main_species_m, species_m)[i]

    qc <- fread(
      paste0(
        "../angsd/results/datasets/historical-",
        f,
        "/qc/historical-",
        f,
        ".",
        g,
        "_all.sampleqc.tsv"
      ),
      header = TRUE,
      sep = "\t"
    )
    qc <- qc[, c("sample", "allsites-filts.depth.mean")]
    hz <- fread(
      paste0(
        "../angsd/results/datasets/historical-",
        f,
        "/analyses/heterozygosity/historical-",
        f,
        ".",
        g,
        "_all.dp6_allsites-filts.filtered_mindp5-allsites-jointcall_allbal0.21-0.79.notrans.fmiss",
        m,
        ".bcftools.het.tsv"
      ),
      header = TRUE,
      sep = "\t"
    )
    samples <- fread(
      paste0(
        "../angsd/results/datasets/historical-",
        f,
        "/poplists/historical-",
        f,
        "_all.indiv.list"
      ),
      header = TRUE,
      sep = "\t",
      select = c("sample", "population", "time", "depth")
    )
    hz <- merge(hz, samples, by = "sample")
    hz <- merge(hz, qc, by = "sample")
    if (s %in% species) {
      hz$year <- str_sub(hz$population, 1, 4)
      hz$region <- "one_region"
    } else {
      hz$year <- str_sub(hz$population, -4, -1)
      hz$region <- str_sub(hz$population, 1, -5)
      # Lump SW Skåne with W Skåne for Pl. argus
      hz[hz$region == "SWSkane", ]$region <- "WSkane"
    }
    if (nrow(hz) > 0) {
      regions <- unique(hz$region)

      relhet <- c()

      for (r in regions) {
        if (r == "SESkane") {
          historical <- hz[hz$region == "ESkane" & hz$year < 1995, ]$sample
        } else {
          historical <- hz[hz$region == r & hz$year < 1995, ]$sample
        }
        modern <- hz[hz$region == r & hz$year > 1995, ]$sample
        if (length(historical) > 0 & length(modern) > 0) {
          combos <- expand.grid(historical, modern)
          colnames(combos) <- c("historical", "modern")
          combos <- merge(
            combos,
            hz[, c("sample", "hetperbp", "year")],
            by.x = "historical", by.y = "sample"
          )
          colnames(combos) <- c("historical", "modern", "histhz", "histyear")
          combos <- merge(
            combos,
            hz[, c("sample", "hetperbp", "year")],
            by.x = "modern", by.y = "sample"
          )
          colnames(combos) <- c(
            "historical", "modern", "histhz", "histyear", "modhz",
            "modyear"
          )
          combos$histyear <- as.numeric(combos$histyear)
          combos$modyear <- as.numeric(combos$modyear)
          combos$het_retained <- (combos$modhz / combos$histhz)
          combos$het_retained_peryear <- exp(
            log(combos$het_retained) / (combos$modyear - combos$histyear)
          )
          combos$rel_delta_het <- combos$het_retained - 1
          combos$rel_delta_het_peryear <- combos$het_retained_peryear - 1
          combos$pred_het_ret_100 <- (combos$het_retained_peryear^100) * 100
          if (s %in% c(
            "aagestis", "cminimus", "palcon", "pamandus", "pargyrognomon"
          )) {
            r <- "WSkane"
          } else if (s == "aoptilete") {
            r <- "NSmaland"
          } else if (s == "cargiolus") {
            r <- "Oland"
          }
          combos$region <- r
          combos$species <- s
          relhet_all <- rbind(relhet_all, combos)
          relhet_all$historical <- as.character(relhet_all$historical)
          relhet_all$modern <- as.character(relhet_all$modern)
        } else {
          df <- data.frame(
            "hist", "mod", 0, 0, 0, 0, 0, 0, 0, 0, 0, "other_region", s
          )
          colnames(df) <- colnames(relhet_all)
          relhet_all <- rbind(relhet_all, df)
        }
      }
    } else {
      df <- data.frame(
        "hist", "mod", 0, 0, 0, 0, 0, 0, 0, 0, 0, "other_region", s
      )
      colnames(df) <- colnames(relhet_all)
      relhet_all <- rbind(relhet_all, df)
    }
  }

relhet_meds <- relhet_all %>%
  group_by(species, region) %>%
  summarise(
    med = median(pred_het_ret_100),
    min = min(pred_het_ret_100),
    max = max(pred_het_ret_100)
  )

relhet_meds[ , c(3:5)] <- round(relhet_meds[ , c(3:5)], 2)

write.table(relhet_meds, "../tables/relhet_meds.tsv",
  sep = "\t", quote = FALSE,
  row.names = FALSE
)
```

After calculating the heterozygosity retention per species/population, we can
assemble the results into a figure.

```{r}
#| label: fig-other-blues
#| fig-width: 10
#| fig-height: 4.5

species_names <- c(
  "Aricia agestis", "Agriades optilete", "Cupido minimus",
  #"Celastrina argiolus",
  "Cyaniris semiargus",
  #"Phengaris alcon*", "Polyommatus amandus*",
  "Plebejus argus",
  "Plebejus argyrognomon*", "Polyommatus icarus", "Scolitantides orion*"
)

cons_status <- c(
  "(LC)", "(LC)", "(NT)",
  # "(LC)",
  "(LC)",
  # "(EN)", "(LC)",
  "(LC)",
  "(CR)", "(LC)", "(EN)"
)

full_name <- paste0(species_names, " ", cons_status)

spec_cons <- data.frame(t(rbind(species_names, cons_status, full_name)))
spec_cons$cons_status <- factor(spec_cons$cons_status,
  levels = c("(LC)", "(NT)", "(EN)", "(CR)")
)

relhet_all$species <- factor(relhet_all$species,
  labels = spec_cons$full_name
)
relhet_all$species <- factor(relhet_all$species,
  levels = spec_cons[order(
    spec_cons$cons_status,
    spec_cons$species_name
  ), ]$full_name
)
relhet_all$region <- factor(relhet_all$region,
  levels = c("WSkane", "ESkane", "SESkane", "NSmaland", "Oland", "one_region")
)

colors <- c("#244F6E", "#658399", "#658399", "#8A2A3B", "#fecd03", "black")

rects <- data.frame(
  xmin = -Inf, xmax = Inf,
  ymin = c(0, 75, 95),
  ymax = c(75, 95, Inf),
  fill = factor(c("alarm", "warning", "acceptable"),
    levels = c("alarm", "warning", "acceptable")
  )
)

ggplot(
    relhet_all,
    aes(x = species, y = pred_het_ret_100, color = region)
  ) +
    geom_rect(
      data = rects,
      aes(ymin = ymin, ymax = ymax, xmin = xmin, xmax = xmax, fill = fill),
      inherit.aes = FALSE, alpha = 0.1
    ) +
    scale_fill_manual(values = c("red", "yellow", "green")) +
    scale_color_manual(
      values = colors
    ) +
    geom_hline(yintercept = 75, linetype = "dashed") +
    geom_hline(yintercept = 95, linetype = "dashed") +
    stat_summary(
      fun.data = "median_hilow", geom = "pointrange",
      position = position_dodge(width = 0.5), size = 0.8, linewidth = 1
    ) +
    ylab("Percent retention of heterozygosity over 100 years") +
    scale_y_continuous(limits = c(0, 111), expand = c(0, 0)) +
    scale_x_discrete(labels = ~ str_wrap(.x, 9)) +
    theme_classic() +
    theme(
      axis.text.x = element_text(face = "italic"),
      axis.title.x = element_blank(),
      legend.position = "none"
    )

ggsave("../figures/fig-specdiv.png", dpi = 600, create.dir = TRUE)
ggsave("../figures/fig-specdiv.svg", dpi = 600, create.dir = TRUE)
```