---
title: "Figure 4"
format:
  html:
    code-fold: true
---

The fourth figure looks at several blue species and tries to gauge their level
of genetic diversity loss in the past century. Many of the species only have a
single sample per time point, so we include a comparison of all possible sample
pair comparisons in the focal species to see if we get the correct

```{r}
#| label: load-packages

library(data.table)
library(stringr)
library(ggplot2)
library(dplyr) # For summarize function (check if still in code blocks)

main_species_f <- c("picarus-notrans", "pargus-notrans", "csemiargus-notrans")
# main_species_f <- c("picarus-mapdam", "pargus-mapdam", "csemiargus-mapdam")
main_species <- c("picarus", "pargus", "csemiargus")
main_genomes <- c("ilPolIcar1.1", "ilPleArgu1.3", "ilCyaSemi1.1")

species <- c(
  "aagestis", "aoptilete", "cargiolus", "cminimus", "palcon",
  "pamandus", "pargyrognomon", "parion", "sorion"
)
# species <- c(
#   "aagestis-mapdam", "aoptilete-mapdam", "cargiolus-mapdam", "cminimus-mapdam", "palcon-mapdam",
#   "pamandus-mapdam", "pargyrognomon-mapdam", "parion-mapdam", "sorion-mapdam"
# )
genomes <- c(
  "ilAriAges2.1", "ilCyaSemi1.1", "ilCelArgi3.2", "ilPolIcar1.1",
  "ilPheArio1.1", "ilPolIcar1.1", "ilPleArgu1.3", "ilPheArio1.1",
  "ilGlaAlex1.1"
)
```

## Heterozygosity change per year

One of the metrics suggested for utilization in Sweden is the per year change in
heterozygosity. While the warning categories are suggested only for significant
differences, which we can't really check with single samples per time period, we
use this same calculation as it is a useful visualization.

```{r}
#| label: main-species-het

# As the main species have specific considerations (i.e. multiple sampling
# areas), they'll be processed separately first.

relhet_all <- c()

dp <- c("full", 4, 3, 2)

for (dep in dp) {
  if (dep == "full") {
    depth <- ""
  } else {
    depth <- paste0(".dp", dep)
  }
  for (i in seq_along(c(main_species, species))) {
    s <- c(main_species, species)[i]
    f <- c(main_species_f, species)[i]
    g <- c(main_genomes, genomes)[i]

    qc <- fread(
      paste0(
        "../angsd/results/datasets/historical-",
        f,
        "/qc/historical-",
        f,
        ".",
        g,
        "_all.sampleqc.tsv"
      ),
      header = TRUE,
      sep = "\t"
    )
    qc <- qc[ , c("sample", "allsites-filts.depth.mean")]
    hz <- fread(
      paste0(
        "../angsd/results/datasets/historical-",
        f,
        "/analyses/heterozygosity/historical-",
        f,
        ".",
        g,
        "_all",
        depth,
        "_allsites-filts_heterozygosity.tsv"
      ),
      header = TRUE,
      sep = "\t"
    )
    samples <- fread(
      paste0(
        "../angsd/results/datasets/historical-",
        f,
        "/poplists/historical-",
        f,
        "_all.indiv.list"
      ),
      header = TRUE,
      sep = "\t",
      select = c("sample", "population", "time", "depth")
    )
    hz <- merge(hz, samples, by = "sample")
    hz <- merge(hz, qc, by = "sample")
    if (dep != "full") {
      hz <- hz[hz$`allsites-filts.depth.mean` > (as.numeric(dep)-0.1), ]
    }
    if (s %in% species) {
      hz$year <- str_sub(hz$population, 1, 4)
      hz$region <- "one_region"
    } else {
      hz$year <- str_sub(hz$population, -4, -1)
      hz$region <- str_sub(hz$population, 1, -5)
      # Lump SW Sk책ne with W Sk책ne and SE Sk책ne with E Sk책ne
      hz[hz$region == "SWSkane", ]$region <- "WSkane"
      hz[hz$region == "SESkane", ]$region <- "ESkane"
    }
    if (nrow(hz) > 0) {
      regions <- unique(hz$region)

      relhet <- c()

      for (r in regions) {
        historical <- hz[hz$region == r & hz$year < 1995, ]$sample
        modern <- hz[hz$region == r & hz$year > 1995, ]$sample
        if (length(historical) > 0 & length(modern) > 0) {
          combos <- expand.grid(historical, modern)
          colnames(combos) <- c("historical", "modern")
          combos <- merge(
            combos,
            hz[, c("sample", "heterozygous_per_1000bp", "year")],
            by.x = "historical", by.y = "sample"
          )
          colnames(combos) <- c("historical", "modern", "histhz", "histyear")
          combos <- merge(
            combos,
            hz[, c("sample", "heterozygous_per_1000bp", "year")],
            by.x = "modern", by.y = "sample"
          )
          colnames(combos) <- c(
            "historical", "modern", "histhz", "histyear", "modhz",
            "modyear"
          )
          combos$histyear <- as.numeric(combos$histyear)
          combos$modyear <- as.numeric(combos$modyear)
          combos$het_retained <- (combos$modhz / combos$histhz)
          combos$het_retained_peryear <- exp(
            log(combos$het_retained) / (combos$modyear - combos$histyear)
          )
          combos$rel_delta_het <- combos$het_retained - 1
          combos$rel_delta_het_peryear <- combos$het_retained_peryear - 1
          combos$pred_het_ret_100 <- (combos$het_retained_peryear^100) * 100
          if (s %in% c(
            "aagestis", "cminimus", "palcon", "pamandus", "pargyrognomon"
          )) {
            r <- "WSkane"
          } else if (s == "aoptilete") {
            r <- "NSmaland"
          } else if (s == "cargiolus") {
            r <- "Oland"
          }
          combos$region <- r
          combos$species <- s
          combos$depth <- dep
          relhet_all <- rbind(relhet_all, combos)
          relhet_all$historical <- as.character(relhet_all$historical)
          relhet_all$modern <- as.character(relhet_all$modern)
        } else {
          df <- data.frame(
            "hist", "mod", 0, 0, 0, 0, 0, 0, 0, 0, 0, "other_region", s, dep
            )
          colnames(df) <- colnames(relhet_all)
          relhet_all <- rbind(relhet_all, df)
        }
      }
    } else {
      df <- data.frame(
        "hist", "mod", 0, 0, 0, 0, 0, 0, 0, 0, 0, "other_region", s, dep
        )
      colnames(df) <- colnames(relhet_all)
      relhet_all <- rbind(relhet_all, df)
    }
  }
}

# Now, we drop Phengaris arion, as unfortunately our modern sample has too low
# sequencing depth (~0.2X, it was also a museum sample).
relhet_all <- relhet_all[relhet_all$species != "parion", ]
# relhet_all <- relhet_all[relhet_all$species != "parion-mapdam", ]
```

```{r}
#| label: fig-other-blues
#| fig-width: 10
#| fig-height: 4.5

species_names <- c(
  "Aricia agestis", "Agriades optilete*",
  "Celastrina argiolus", "Cupido minimus*", "Cyaniris semiargus",
  "Phengaris alcon*", "Polyommatus amandus", "Plebejus argus",
  "Plebejus argyrognomon*", "Polyommatus icarus", "Scolitantides orion*"
)

cons_status <- c(
  "(LC)", "(LC)", "(LC)", "(NT)", "(LC)", "(EN)", "(LC)", "(LC)",
  "(CR)", "(LC)", "(EN)"
)

full_name <- paste0(species_names, " ", cons_status)

spec_cons <- data.frame(t(rbind(species_names, cons_status, full_name)))
spec_cons$cons_status <- factor(spec_cons$cons_status,
  levels = c("(LC)", "(NT)", "(EN)", "(CR)")
)

relhet_all$species <- factor(relhet_all$species,
  labels = spec_cons$full_name
)
relhet_all$species <- factor(relhet_all$species,
  levels = spec_cons[order(
    spec_cons$cons_status,
    spec_cons$species_name
  ), ]$full_name
)

rects <- data.frame(
  xmin = -Inf, xmax = Inf,
  ymin = c(0, 75, 95),
  ymax = c(75, 95, Inf),
  fill = factor(c("alarm", "warning", "acceptable"),
    levels = c("alarm", "warning", "acceptable")
  )
)

plots <- list()

for (d in c("full", "4", "3", "2")) {

  plots[[d]] <- ggplot(
    relhet_all[relhet_all$depth == d, ],
    aes(x = species, y = pred_het_ret_100, color = region)
  ) +
    geom_rect(
      data = rects,
      aes(ymin = ymin, ymax = ymax, xmin = xmin, xmax = xmax, fill = fill),
      inherit.aes = FALSE, alpha = 0.1
    ) +
    scale_fill_manual(values = c("red", "yellow", "green")) +
    scale_color_manual(
      values = c("#658399", "#8A2A3B", "#fecd03", "black", "#244F6E")
    ) +
    geom_hline(yintercept = 75, linetype = "dashed") +
    geom_hline(yintercept = 95, linetype = "dashed") +
    stat_summary(
      fun.data = "median_hilow", geom = "pointrange",
      position = position_dodge(width = 0.5), size = 0.8, linewidth = 1
    ) +
    ylab("Percent retention of heterozygosity over 100 years") +
    scale_y_continuous(expand = c(0, 0)) +
    scale_x_discrete(labels = ~ str_wrap(.x, 9)) +
    theme_classic() +
    theme(
      axis.text.x = element_text(face = "italic"),
      axis.title.x = element_blank(),
      legend.position = "none"
    )
}

plots[["full"]]

ggsave("../figures/fig-specdiv.png", dpi = 600, create.dir = TRUE)
```

```{r}
#| label: suppfig-other-blues
#| fig-width: 10
#| fig-height: 13.5
library(cowplot)

plot_grid(
  plots[["4"]],
  plots[["3"]],
  plots[["2"]],
  ncol = 1,
  labels = c("A", "B", "C")
)

ggsave("../figures/suppfig-specdiv.png", dpi = 600, create.dir = TRUE)
```

```{r}
#| evaluate: FALSE

spmeans <- relhet_all %>%
  group_by(species) %>%
  summarize(modhz = median_hilow(modhz),
            pred_het_ret_100 = median_hilow(pred_het_ret_100))

spmeans <- data.frame(spmeans)

ggplot(spmeans, aes(x = modhz$y, y = pred_het_ret_100$y)) +
  geom_smooth(method = "lm") +
  geom_errorbarh(aes(xmin=modhz$ymin, xmax=modhz$ymax, color = species)) + 
  geom_errorbar(aes(ymin=pred_het_ret_100$ymin, ymax=pred_het_ret_100$ymax, color = species)) +
  geom_point(size = 3, aes(color = species))

summary(lm(spmeans$pred_het_ret_100$y ~ spmeans$modhz$y))

```