---
title: "Quality checking"
format:
  html:
    code-fold: true
---

First, we will take a look at the general quality of our results to make
decisions about final methods, as well as if there are any biases that must be
considered in our analysis.

```{r}
#| label: load-packages
#| code-summary: Load up required packages and setup species and genome lists

library(data.table) # For reading in data while skipping columns
# library(stringr) # For separating pop names into regions and sampling years
library(ggplot2) # For plotting
library(ggpubr) # For adding stats to graphs
library(cowplot) # For arranging plots into a multi-panel figure
# library(ggtext) # For more control over axis text layout
# library(dplyr) # For summarize function and piping
# library(geosphere) # For including distances on F~ST~ plot
# library(scales) # For some control of plot scales
# library(png) # For reading in maps made in QGIS
# library(grid) # For plotting maps made in QGIS with ggplot

species <- c("picarus", "pargus", "csemiargus")
genomes <- c("ilPolIcar1.1", "ilPleArgu1.3", "ilCyaSemi1.1")
```

## Reference bias and DNA damage

As historical samples will have degraded DNA, we need to consider how we will
address this difference between historical and modern samples. One method would
be to rescale the putatively damaged bases with MapDamage2 [@jónsson2013],
another would be to remove transitions, which are what DNA damage is likely to
manifest as. The former will leave us with the most data, but may lead to
overcorrecting true transitions, while the latter should remove damage by
removing all transition, but will remove a lot of variation data.

We can assess how each of these methods perform by comparing the identity by
state similarity to the reference when using each. Additionally, this will allow
us to see if there is a reference bias induced by the shorter fragment length of
the historical samples. All of the following checks will take place using our
filtered sites list, as this removes variation induced by poor mapping around
repetitive content and similar challenging regions.

```{r}
#| label: fig-ibsref
#| fig-width: 7
#| fig-height: 8

ibs_all_trans <- c()

for (i in c(1:length(species))) {
  s <- species[i]
  g <- genomes[i]
  ibs <- fread(
    paste0(
      "../angsd/results/datasets/historical-",
      s,
      "-trans/qc/ibs_refbias/historical-",
      s,
      "-trans.",
      g,
      "_all_allsites-filts.refibs.tsv"
    ),
    header = TRUE,
    sep = "\t",
    select = c("sample", "ibs.to.ref", "population", "time", "depth")
  )
  ibs$species <- s
  ibs_all_trans <- rbind(ibs_all_trans, ibs)
}

ibs_all_trans$species <- factor(ibs_all_trans$species, levels = species)

ibs_trans_plot <- ggplot(ibs_all_trans, aes(x = time, y = ibs.to.ref)) +
  ggtitle("Include Transitions") +
  facet_grid(cols = vars(species)) +
  ylab("Mean IBS Similarity to Reference") +
  geom_boxplot() +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    axis.title.x = element_blank()
  )

ibs_all_notrans <- c()

for (i in c(1:length(species))) {
  s <- species[i]
  g <- genomes[i]
  ibs <- fread(
    paste0(
      "../angsd/results/datasets/historical-",
      s,
      "-notrans/qc/ibs_refbias/historical-",
      s,
      "-notrans.",
      g,
      "_all_allsites-filts.refibs.tsv"
    ),
    header = TRUE,
    sep = "\t",
    select = c("sample", "ibs.to.ref", "population", "time", "depth")
  )
  ibs$species <- s
  ibs_all_notrans <- rbind(ibs_all_notrans, ibs)
}

ibs_all_notrans$species <- factor(ibs_all_notrans$species, levels = species)

ibs_notrans_plot <- ggplot(ibs_all_notrans, aes(x = time, y = ibs.to.ref)) +
  ggtitle("Exclude Transitions (Method used in manuscript)") +
  facet_grid(cols = vars(species)) +
  ylab("Mean IBS Similarity to Reference") +
  geom_boxplot() +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    axis.title.x = element_blank()
  )

ibs_all_mapdam <- c()

for (i in c(1:length(species))) {
  s <- species[i]
  g <- genomes[i]
  ibs <- fread(
    paste0(
      "../angsd/results/datasets/historical-",
      s,
      "-mapdam/qc/ibs_refbias/historical-",
      s,
      "-mapdam.",
      g,
      "_all_allsites-filts.refibs.tsv"
    ),
    header = TRUE,
    sep = "\t",
    select = c("sample", "ibs.to.ref", "population", "time", "depth")
  )
  ibs$species <- s
  ibs_all_mapdam <- rbind(ibs_all_mapdam, ibs)
}


ibs_all_mapdam$species <- factor(ibs_all_mapdam$species, levels = species)

ibs_mapdam_plot <- ggplot(ibs_all_mapdam, aes(x = time, y = ibs.to.ref)) +
  ggtitle("MapDamage Rescaling") +
  facet_grid(cols = vars(species)) +
  ylab("Mean IBS Similarity to Reference") +
  geom_boxplot() +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    axis.title.x = element_blank()
  )

plot_grid(
  ibs_trans_plot,
  ibs_mapdam_plot,
  ibs_notrans_plot,
  nrow = 3,
  labels = c("A", "B", "C")
)

ggsave("../figures/suppfig-ibs.png")
```

We can see that the uncorrected historical samples are more distant to the
reference than the modern ones, and that there is more variation in the
individual IBS scores in the historical samples. This is good, as it suggests
that our samples do not have a reference bias from smaller fragment size.
Likely, the extra variation in IBS score is due to differences in per sample DNA
damage, and we can see that while both methods of correction reduce this
variation, it is best in the transition removal scores. Additionally, while in
*C. semiargus*, both methods do not 'overcorrect' and result in reference
biased historical samples, in the other two species, MapDamage induces
substantial reference bias to the historical samples.

We can confirm that this is likely a bias and not a true shorter distance to the
reference for historical samples in these species by looking at the
transition/transversion ratios for genotype calls:

As we can see, when performing genotype calling, we get lower
transition/transversion ratios in historical samples compared to a consistent
uniform rate in the modern ones. This suggests that transition reads that would
support likely 'true' transitions are being filtered too much due to the
MapDamage rescaling.

For this reason, we decide to move forward using transition removal as our DNA
damage correction method. Next, we check to make sure that after this correction
we don't see any biases from fragment length or depth of coverage in sample IBS
scores.

```{r}
#| label: fig-covlenibs
#| eval: false

mapdam_all <- data.frame()

histsamp <- ibs_all_notrans[ibs_all_notrans$time == "historical", ]$sample

for (i in histsamp) {
  s <- ibs_all_notrans[ibs_all_notrans$sample == i, ]$species
  g <- genomes[match(s, species)]
  dam <- read.table(
    paste0(
      "../angsd/results/mapping/qc/damageprofiler/",
      i,
      ".",
      g,
      "/5pCtoT_freq.txt"
    ),
    header = TRUE,
    sep = "\t",
    comment.char = "#"
  )
  avgdam <- mean(dam[c(1:5), "X5pC.T"])
  lgdist <- read.table(
    paste0(
      "../angsd/results/mapping/qc/damageprofiler/",
      i,
      ".",
      g,
      "/lgdistribution.txt"
    ),
    header = TRUE,
    sep = "\t",
    comment.char = "#"
  )
  lgdist <- lgdist[lgdist$Std == "+", ]
  medlen <- with(lgdist, median(rep(x = Length, times = Occurrences)))
  row <- c(i, avgdam, medlen, species[s])
  mapdam_all <- unname(mapdam_all)
  mapdam_all <- rbind(mapdam_all, row)
}

colnames(mapdam_all) <- c("sample", "avgdam", "medlen", "species")

qc_all <- c()

for (s in species) {
  g <- genomes[match(s, species)]
  for (m in c("mapdam", "notrans")) {
    qc <- fread(
      paste0(
        "../angsd/results/datasets/historical-",
        s,
        "-",
        m,
        "/qc/historical-",
        s,
        "-",
        m,
        ".",
        g,
        "_all.sampleqc.tsv"
      ),
      header = TRUE,
      sep = "\t",
      select = c("sample", "allsites-filts.depth.mean")
    )
    qc_all <- rbind(qc_all, qc)
  }
}

mapdam_mapdam <- merge(mapdam_all, ibs_all_mapdam, by = c("sample", "species"))
mapdam_notrans <- merge(mapdam_all, ibs_all_notrans, by = c("sample", "species"))
mapdam_mapdam <- merge(mapdam_mapdam, qc_all, by = "sample")
mapdam_notrans <- merge(mapdam_notrans, qc_all, by = "sample")

ibs_cov_all <- merge(ibs_all_notrans, qc_all, by = "sample")

ggplot(mapdam_mapdam, aes(x = medlen, y = ibs.to.ref)) +
  facet_grid(cols = vars(species), scales = "free") +
  geom_point() +
  geom_smooth(method = "lm") +
  stat_cor()

ggplot(mapdam_notrans, aes(x = `allsites-filts.depth.mean`, y = ibs.to.ref)) +
  facet_grid(cols = vars(species), scales = "free") +
  geom_point() +
  geom_smooth(method = "lm") +
  stat_cor(label.y = 0.99) +
  stat_regline_equation(label.y = 0.989)
```

It appears that while fragment size has no notable impact on IBS scores, mean
sequencing depth has a slightly negative relationship with IBS scores in all
three species. However, the difference in IBS scores from the highest and lowest
depth historical samples are relatively small, 0.0021 in *P. icarus*, 0.0012 in
*P. argus*, and 0.0006 in *C. semiargus*, so likely there will be little impact
on the results, but we will have to keep this in mind and downsample sequencing
depth to confirm our results are consistent after controlling for this.